```{r}
#| include: false
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

library(dplyr)
library(ggplot2)
library(glmmTMB)
library(texreg)
library(ggeffects)
library(patchwork)
library(gt)
library(kableExtra)
library(here)
library(margins)
library(broom)
library(modelsummary)
library(stargazer)
library(tidyr)

# Load data
load(here("input/data/proc/elsoc_final_2.RData"))
```

# Results

This section presents the findings on the factors shaping the justification of violence in the context of social protest in Chile.

```{r}
#| results: false


# 1st Step: Convert to non-ordered factor
elsoc_final_2 <- elsoc_final_2 %>%
  mutate(
    educ_cat_unordered = factor(educ_cat_factor, ordered = FALSE),
    movilidad_cat_unordered = factor(movilidad_cat_factor, ordered = FALSE)
  )

# 2nd Step: Create consistent index of violence in protests (only variables in all waves)
elsoc_final_2 <- elsoc_final_2 %>%
  mutate(
    # Consistent temporal index (students and workers only)
    justif_violencia_protesta_consistente = rowMeans(
      dplyr::select(., violencia_trabajadores, violencia_estudiantes),
      na.rm = TRUE
    ),
    # Broad index (all variables available per wave)
    justif_violencia_protesta_amplio = justif_violencia_protesta
  )

# Check availability per wave
elsoc_final_2 %>%
  group_by(year) %>%
  summarise(
    n = n(),
    estudiantes = sum(!is.na(violencia_estudiantes)),
    trabajadores = sum(!is.na(violencia_trabajadores)),
    inmobiliario = sum(!is.na(violencia_inmobiliario)),
    transporte = sum(!is.na(violencia_transporte)),
    locales = sum(!is.na(violencia_locales)),
    carab_marchas = sum(!is.na(violencia_carabineros_marchas)),
    carab_tomas = sum(!is.na(violencia_carabineros_tomas))
  ) %>%
  print()

```

## Descriptive Statistics

### Justification of Violence by Education Level

The patterns observed in @tbl-justificacion-educ reveal a notable trend. Among non-participants, higher education levels tend to be associated with lower justification of protest violence, with means declining from approximately 1.90 to 1.80. However, this trend reverses among those who do participate: participants with university education display the highest justification levels, reaching means around 2.29 and 2.27. This differentiated effect of education appears to be specific to protest violence, as, by contrast, the justification of state violence shows a much more stable pattern, with considerably smaller variation across educational levels.

```{r}
#| label: tbl-justificacion-educ
#| tbl-cap: "Justification of violence by education level and protest participation"
#| echo: false
#| warning: false
#| results: asis

# Redefine factor labels:
elsoc_final_2 <- elsoc_final_2 %>%
  mutate(
    educ_cat_unordered = factor(
      educ_cat_unordered,
      labels = c(
        "High School or less",
        "Technical College incomplete",
        "Technical College complete",
        "University incomplete",
        "University complete"
      )
    )
  )

# Create and display table
tabla_1 <- elsoc_final_2 %>%
  filter(!is.na(educ_cat_unordered) & !is.na(protesta_dummy)) %>%
  group_by(educ_cat_unordered, Participation = factor(protesta_dummy,
                                                       labels = c("Yes", "No"))) %>%
  summarise(
    N = n(),
    `Protest Violence` = round(mean(justif_violencia_protesta_consistente, na.rm = TRUE), 2),
    `SD` = round(sd(justif_violencia_protesta_consistente, na.rm = TRUE), 2),
    `State Violence` = round(mean(justif_violencia_estatal, na.rm = TRUE), 2),
    `SD ` = round(sd(justif_violencia_estatal, na.rm = TRUE), 2),
    .groups = "drop"
  ) %>%
  knitr::kable(
    col.names = c("Education", "Participation", "N", "Mean", "SD", "Mean", "SD"),
    align = c("l", "l", "c", "c", "c", "c", "c"),
    format = "latex",
    booktabs = TRUE
  ) %>%
  kable_styling(
    latex_options = c("striped", "hold_position", "scale_down"),
    font_size = 9,
    position = "center"
  ) %>%
  add_header_above(c(" " = 3, "Protest Violence" = 2, "State Violence" = 2)) %>%
  footnote(
    general = "Compiled by the author based on ELSOC (2016–2023). Scale of 1–5, where 1 = Never justified and 5 = Always justified",
    threeparttable = TRUE
  )

tabla_1
```


### Temporal Evolution: Justification of Violence and Protest Participation (2016–2023)

```{r}
#| label: fig-evolucion-temporal
#| fig-cap: "Evolution of protest participation 2016–2023"
#| fig-width: 12
#| fig-height: 7
#| echo: false

# Participation Panel
plot_evol_particip <- elsoc_final_2 %>%
  group_by(year) %>%
  summarise(
    Participation = 100 * mean(protesta_dummy, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = year, y = Participation)) +
  geom_vline(xintercept = 2019, linetype = "dashed", color = "red", alpha = 0.6) +
  geom_line(linewidth = 1.2, color = "#619CFF") +
  geom_point(size = 2.5, color = "#619CFF") +
  labs(
    title = "Protest Participation Over Time",
    x = "Year",
    y = "% who Participated") +
  theme_minimal(base_size = 14)

# Combine
grafico_evolucion <- plot_evol_particip +
  plot_annotation(
    tag_levels = 'A',
    caption = "Compiled by the author based on ELSOC 2016-2023. Note: 2019 = Social Outbreak",
    theme = theme(plot.caption = element_text(size = 11))
  )

# Display plot
grafico_evolucion
```

The temporal pattern analysis of @fig-evolucion-temporal shows a result consistent with the observed protest cycle context. In 2019 — the year of the social outbreak — the highest average protest participation rate is recorded, reflecting the intensity of mobilization. In the post-2019 period, although participation declines from its peak, it remains at slightly higher levels than those observed before 2019, until its fall between 2021 and 2023, where it reaches its lowest average across the entire reported cycle.

```{r}
#| label: fig-participacion-estratificada
#| fig-cap: "Temporal divergence between protest violence and state violence justification (2016–2023)"
#| fig-width: 12
#| fig-height: 18
#| echo: false
#| warning: false

# Panel: Direct comparison Protest vs State violence by year
comparacion_data <- elsoc_final_2 %>%
  group_by(year) %>%
  summarise(
    `Protest Violence` = mean(justif_violencia_protesta_consistente, na.rm = TRUE),
    `State Violence` = mean(justif_violencia_estatal, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  tidyr::pivot_longer(cols = c("Protest Violence", "State Violence"),
                      names_to = "Type",
                      values_to = "Justification")

just_viol <- ggplot(comparacion_data, aes(x = year, y = Justification, color = Type, group = Type)) +
  geom_line(aes(linetype = Type), linewidth = 1.5) +
  geom_point(aes(shape = Type), size = 3.5) +
  geom_vline(xintercept = 2019, linetype = "dashed", color = "red", alpha = 0.6) +
  scale_color_manual(values = c("Protest Violence" = "#EFC000FF",
                                  "State Violence" = "#0073C2FF"),
                     name = "Type of Violence") +
  scale_linetype_manual(values = c("Protest Violence" = "solid",
                                    "State Violence" = "dashed"),
                        name = "Type of Violence") +
  scale_shape_manual(values = c("Protest Violence" = 16,
                                 "State Violence" = 17),
                     name = "Type of Violence") +
  labs(
    title = "Divergence between Justification of Violence in Protests and State Violence",
    x = "Year",
    y = "Justification (scale 1-5)",
    caption = "Red line: 2019 social outbreak. Compiled by the author based on ELSOC 2016–2023."
  ) +
  theme_minimal(base_size = 15) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 11),
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 11, color = "gray30"),
    plot.caption = element_text(size = 10, hjust = 0)
  ) +
  ylim(1.4, 2.1)

# Display plot
just_viol
```

@fig-participacion-estratificada reveals a striking temporal inversion in the moral hierarchy surrounding political violence. Pre-2019 (2016–2018), state violence consistently received higher justification than protest violence at the aggregate level (gap of −0.09 to −0.26 points), suggesting a baseline in which coercive force by state agents was viewed more leniently than disruptive actions by protesters. In 2019, this hierarchy reverses dramatically: protest violence rises to its sample-wide peak (~1.91) while state violence falls to its minimum (~1.53), producing a gap in the opposite direction (+0.38 points, protest > state). This reversal reflects the impact of the uprising — both the surge in protest participation and the delegitimization of state repression in a context of widely perceived police brutality.

In the post-2019 period (2022–2023), the pre-2019 hierarchy reasserts itself: state violence justification rises progressively while protest violence declines, with state again exceeding protest justification by 0.15–0.29 points. This pattern suggests that 2019 constituted a critical but temporary moral inversion — the uprising briefly reversed the default hierarchy, but the post-uprising period has seen a reversion in the broader population toward the baseline pattern in which state coercion is viewed more permissively than collective disruptive action. The aggregate convergence therefore reflects the attitudinal trajectories of the non-participating majority more than those of habitual protesters, whose patterns are captured in the interaction models that follow.

```{r}
#| include: false

# Function to calculate means and CIs by group and period (pre/post 2019)
calc_means_ci <- function(data, group_var, tipo_violencia) {
  data %>%
    filter(!is.na(year) & year != 2019 & !is.na(!!sym(group_var)) & !is.na(protesta_dummy)) %>%
    mutate(period = factor(
      ifelse(year < 2019, "Pre-2019\n(2016–2018)", "Post-2019\n(2022–2023)"),
      levels = c("Pre-2019\n(2016–2018)", "Post-2019\n(2022–2023)")
    )) %>%
    group_by(period, !!sym(group_var), protesta_dummy) %>%
    summarise(
      mean = mean(!!sym(tipo_violencia), na.rm = TRUE),
      se = sd(!!sym(tipo_violencia), na.rm = TRUE) / sqrt(n()),
      n = n(),
      .groups = "drop"
    ) %>%
    mutate(
      ci_low = mean - 1.96 * se,
      ci_high = mean + 1.96 * se,
      participation = factor(protesta_dummy,
                            levels = c(0, 1),
                            labels = c("Did not participate", "Participated"))
    )
}

# Panel Education × Protest Violence
data_educ_prot <- calc_means_ci(elsoc_final_2, "educ_cat_unordered", "justif_violencia_protesta_consistente")

p1 <- ggplot(data_educ_prot, aes(x = educ_cat_unordered, y = mean, color = participation, group = participation)) +
  geom_point(aes(shape = participation), position = position_dodge(width = 0.5), size = 2.5) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high),
                position = position_dodge(width = 0.5), width = 0.2, linewidth = 0.8) +
  geom_line(aes(linetype = participation), position = position_dodge(width = 0.5), linewidth = 1) +
  facet_wrap(~period, ncol = 2) +
  scale_color_manual(values = c("Did not participate" = "#F8766D", "Participated" = "#00BFC4"), name = "") +
  scale_linetype_manual(values = c("Did not participate" = "dashed", "Participated" = "solid"), name = "") +
  scale_shape_manual(values = c("Did not participate" = 17, "Participated" = 16), name = "") +
  labs(title = "Protest Violence",
       x = "Education Level",
       y = "Justification (scale 1-5)") +
  theme_minimal(base_size = 15) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        legend.position = "bottom",
        legend.text = element_text(size = 11),
        strip.text = element_text(face = "bold", size = 12)) +
  ylim(1, 3)

# Panel Education × State Violence
data_educ_est <- calc_means_ci(elsoc_final_2, "educ_cat_unordered", "justif_violencia_estatal")

p2 <- ggplot(data_educ_est, aes(x = educ_cat_unordered, y = mean, color = participation, group = participation)) +
  geom_point(aes(shape = participation), position = position_dodge(width = 0.5), size = 2.5) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high),
                position = position_dodge(width = 0.5), width = 0.2, linewidth = 0.8) +
  geom_line(aes(linetype = participation), position = position_dodge(width = 0.5), linewidth = 1) +
  facet_wrap(~period, ncol = 2) +
  scale_color_manual(values = c("Did not participate" = "#F8766D", "Participated" = "#00BFC4"), name = "") +
  scale_linetype_manual(values = c("Did not participate" = "dashed", "Participated" = "solid"), name = "") +
  scale_shape_manual(values = c("Did not participate" = 17, "Participated" = 16), name = "") +
  labs(title = "State Violence",
       x = "Education Level",
       y = "Justification (scale 1-5)") +
  theme_minimal(base_size = 15) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        legend.position = "bottom",
        legend.text = element_text(size = 11),
        strip.text = element_text(face = "bold", size = 12)) +
  ylim(1, 3)

# Combine the 2 panels by education
educ_plot <- (p1 / p2) +
  plot_layout(guides = "collect") +
  plot_annotation(
    tag_levels = 'A',
    caption = "Points show means; bars show 95% CI. Periods aggregate 2016–2018 (pre) and 2022–2023 (post). Year 2019 excluded. Compiled by the author based on ELSOC."
  ) &
  theme(
    legend.position = "bottom",
    plot.caption = element_text(size = 7)
  )
```


```{r}
#| label: fig-estallido-1
#| fig-cap: "Comparison of violence justification before (2016–2018) and after (2022–2023) the social outbreak by education and participation"
#| fig-width: 12
#| fig-height: 12


# Display plot
educ_plot
```

@fig-estallido-1 presents the direct comparison of violence justification by education and participation status between the pre-2019 period (2016–2018) and the post-2019 period (2022–2023), providing evidence on the permanent character of the political shock.

Panel A (protest violence) shows that among non-participants all educational groups register a moderate increase post-2019, while among participants all levels maintain consistently elevated justifications (~2.0–2.3) with overlapping confidence intervals — mobilization experience equalizes attitudes regardless of cultural capital, and the gap between participants and non-participants remains stable across both periods, consistent with the participation effect persisting years after the outbreak. Panel B (state violence) shows a more dramatic shift: among non-participants a substantial increase is observed post-2019 across all educational groups, reflecting the normalization of repression justification even among more educated sectors. Among participants, the pattern reverses — all groups show lower justifications post-2019, with university completers maintaining consistently low levels throughout. The pre/post contrast is consistent with the 2019 outbreak having generated a persistent inflection point, with divergent trajectories apparent 3–4 years after the critical event.


## Regression Analyses

```{r}
#| echo: false
#| include: false

# ============================================
# EDUCATION MODELS
# ============================================

# Model 1a: Education - Main effects for Protest Violence
mod_educ_protesta_main <- glmmTMB(
  justif_violencia_protesta_consistente ~ educ_cat_unordered + protesta_dummy +
    edad + mujer + ideologia_std + factor(year) +
    (1 | idencuesta),
  data = elsoc_final_2,
  family = gaussian()
)

# Model 1b: Education × Protest for Protest Violence (with interaction)
mod_educ_protesta_int <- glmmTMB(
  justif_violencia_protesta_consistente ~ educ_cat_unordered * protesta_dummy +
    edad + mujer + ideologia_std + factor(year) +
    (1 | idencuesta),
  data = elsoc_final_2,
  family = gaussian()
)

# Model 2a: Education - Main effects for State Violence
mod_educ_estatal_main <- glmmTMB(
  justif_violencia_estatal ~ educ_cat_unordered + protesta_dummy +
    edad + mujer + ideologia_std + factor(year) +
    (1 | idencuesta),
  data = elsoc_final_2,
  family = gaussian()
)

# Model 2b: Education × Protest for State Violence (with interaction)
mod_educ_estatal_int <- glmmTMB(
  justif_violencia_estatal ~ educ_cat_unordered * protesta_dummy +
    edad + mujer + ideologia_std + factor(year) +
    (1 | idencuesta),
  data = elsoc_final_2,
  family = gaussian()
)
```

```{r}
#| label: tbl-modelos-principales
#| tbl-cap: "Main models of violence justification"
#| echo: false
#| warning: false
#| results: asis

# Custom coefficient map
ccoef <- list(
  "educ_cat_unorderedTechnical College incomplete" = "Technical College incomplete",
  "educ_cat_unorderedTechnical College complete" = "Technical College complete",
  "educ_cat_unorderedUniversity incomplete" = "University incomplete",
  "educ_cat_unorderedUniversity complete" = "University complete",
  "protesta_dummy" = "Protest participation",
  "edad" = "Age",
  "mujer" = "Woman",
  "ideologia_std" = "Ideology",
  "factor(year)2017" = "2017",
  "factor(year)2018" = "2018",
  "factor(year)2019" = "2019",
  "factor(year)2022" = "2022",
  "factor(year)2023" = "2023"
)

{
  .tex <- capture.output(
    texreg(list(mod_educ_protesta_main, mod_educ_estatal_main),
       custom.model.names = c("Model 1", "Model 2"),
       stars = c(0.001, 0.01, 0.05),
       omit.coef = "(Intercept)",
       custom.coef.map = ccoef,
       digits = 3,
       groups = list("Education (Ref. = High School or less)" = 1:4,
         "Controls" = 5:8,
         "Year (Ref. = 2016)" = 9:13),
       custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05",
       leading.zero = FALSE,
       use.packages = FALSE,
       booktabs = TRUE,
       dcolumn = FALSE,
       scalebox = 0.8,
       include.bic = FALSE,
       include.loglik = FALSE,
       include.aic = TRUE,
       center = TRUE,
       custom.gof.names = c("AIC", "Num. obs.", "Num. groups: idencuesta", "Var: idencuesta"),
       reorder.gof = c(3, 4, 2, 1, 5),
       custom.gof.rows = list("BIC" = c(BIC(mod_educ_protesta_main), BIC(mod_educ_estatal_main))),
       caption = NULL,
       table.placement = "",
       label = "",
       float.pos = "!htbp")
  )
  cat(.tex[!grepl("\\begin{table",  .tex, fixed = TRUE) &
           !grepl("\\end{table}",   .tex, fixed = TRUE) &
           !grepl("\\begin{center}", .tex, fixed = TRUE) &
           !grepl("\\end{center}",  .tex, fixed = TRUE)],
      sep = "\n")
}
```


@tbl-modelos-principales presents two main models guiding the analysis of this study. Model 1 examines the effects of education and protest participation on protest violence justification, and Model 2 evaluates those same effects on state violence justification. Both models include random effects by individual and controls for age, gender, ideology, and year.

The results reveal differentiated patterns depending on the type of violence. For protest violence, participation shows a consistent positive effect, increasing justification by approximately 0.36 points. By contrast, for state violence, the direct effect of participation is negative and highly significant, showing that protest participants do not more readily justify police repression.

The controls operate consistently: right-wing ideology reduces justification of protest violence (~-0.03 per point on the scale) but increases it for state violence (~+0.05). Women reject state violence more (~-0.13) with no differences on protest violence. The year 2019 shows the most dramatic contrast: a moderate increase in protest violence justification but a sharp fall in state violence justification (~-0.40), reflecting the impact of the social outbreak on attitudes toward police repression.

```{r}
#| label: tbl-modelos-interaccion
#| tbl-cap: "Interaction models for violence justification"
#| echo: false
#| warning: false
#| results: asis

# Custom coefficient map for interaction models
ccoef_int <- list(
  "educ_cat_unorderedTechnical College incomplete" = "Technical College incomplete",
  "educ_cat_unorderedTechnical College complete" = "Technical College complete",
  "educ_cat_unorderedUniversity incomplete" = "University incomplete",
  "educ_cat_unorderedUniversity complete" = "University complete",
  "protesta_dummy" = "Protest participation",
  "educ_cat_unorderedTechnical College incomplete:protesta_dummy" = "Technical College incomplete × Protest",
  "educ_cat_unorderedTechnical College complete:protesta_dummy" = "Technical College complete × Protest",
  "educ_cat_unorderedUniversity incomplete:protesta_dummy" = "University incomplete × Protest",
  "educ_cat_unorderedUniversity complete:protesta_dummy" = "University complete × Protest",
  "edad" = "Age",
  "mujer" = "Woman",
  "ideologia_std" = "Ideology",
  "factor(year)2017" = "2017",
  "factor(year)2018" = "2018",
  "factor(year)2019" = "2019",
  "factor(year)2022" = "2022",
  "factor(year)2023" = "2023"
)

{
  .tex <- capture.output(
    texreg(list(mod_educ_protesta_int, mod_educ_estatal_int),
       custom.model.names = c("Model 1b", "Model 2b"),
       stars = c(0.001, 0.01, 0.05),
       omit.coef = "(Intercept)",
       custom.coef.map = ccoef_int,
       digits = 3,
       groups = list("Education (Ref. = High School or less)" = 1:4,
         "Education Interactions" = 6:9,
         "Controls" = 10:12,
         "Year (Ref. = 2016)" = 13:17),
       custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05",
       leading.zero = FALSE,
       use.packages = FALSE,
       booktabs = TRUE,
       dcolumn = FALSE,
       scalebox = 0.8,
       include.bic = FALSE,
       include.loglik = FALSE,
       include.aic = TRUE,
       center = TRUE,
       custom.gof.names = c("AIC", "Num. obs.", "Num. groups: idencuesta", "Var: idencuesta"),
       reorder.gof = c(3, 4, 2, 1, 5),
       custom.gof.rows = list("BIC" = c(BIC(mod_educ_protesta_int), BIC(mod_educ_estatal_int))),
       caption = NULL,
       table.placement = "",
       label = "",
       float.pos = "!htbp")
  )
  cat(.tex[!grepl("\\begin{table",  .tex, fixed = TRUE) &
           !grepl("\\end{table}",   .tex, fixed = TRUE) &
           !grepl("\\begin{center}", .tex, fixed = TRUE) &
           !grepl("\\end{center}",  .tex, fixed = TRUE)],
      sep = "\n")
}
```

The models in @tbl-modelos-interaccion reveal that the effect of participation varies significantly by education level. For protest violence, the interactions between education and participation are positive, indicating that the legitimizing effect of participating is stronger among more educated individuals. A person with complete university education who participates increases their justification by 0.40 points versus 0.27 for secondary education, evidencing the "educational paradox": education reduces justification among non-participants (-0.13***) but amplifies the effect of participation.

For state violence, the pattern reverses completely. The interactions are negative (university incomplete: -0.16*; university complete: -0.18***), indicating that only the combination of university education + participation generates significant rejection of repression (total effect: -0.22 points). Education alone (+0.06, ns) or participation alone (-0.04, ns) are insufficient; it is their interaction that activates the differentiated rejection.

The bidirectional reconfiguration (legitimizing protests, rejecting repression) is clearest among university-educated participants, consistent with the dual moral framework requiring both higher education and direct mobilization experience.

### Disentangling Selection from Socialization: A Within-Between Analysis

A key interpretive challenge in the models above is that the participation effect may reflect selection (individuals with pre-existing radical attitudes self-select into protests) rather than socialization (participation itself transforms attitudes). The random-intercept models above control for stable individual differences but do not directly decompose the participation effect into these two components. To address this, we estimate a within-between (REWB) model [@bell_explaining_2015a] that decomposes protest participation into: (a) the within-person component — deviations from each individual's mean participation across waves, capturing attitudinal change associated with varying participation over time — and (b) the between-person component — each individual's mean participation, capturing stable propensity to participate and thereby indexing potential selection bias.

```{r}
#| include: false

# ============================================================
# REWB: Within-between decomposition of participation effect
# ============================================================

elsoc_final_2 <- elsoc_final_2 %>%
  group_by(idencuesta) %>%
  mutate(
    protesta_mean   = mean(protesta_dummy, na.rm = TRUE),
    protesta_within = protesta_dummy - protesta_mean
  ) %>%
  ungroup()

# REWB Model 3a: Protest violence
mod_rewb_protesta <- glmmTMB(
  justif_violencia_protesta_consistente ~
    educ_cat_unordered +
    protesta_within +
    protesta_mean +
    edad + mujer + ideologia_std + factor(year) +
    (1 | idencuesta),
  data = elsoc_final_2,
  family = gaussian()
)

# REWB Model 3b: State violence
mod_rewb_estatal <- glmmTMB(
  justif_violencia_estatal ~
    educ_cat_unordered +
    protesta_within +
    protesta_mean +
    edad + mujer + ideologia_std + factor(year) +
    (1 | idencuesta),
  data = elsoc_final_2,
  family = gaussian()
)

# REWB Interaction: Education × within-person and between-person decomposition
mod_rewb_int_protesta <- glmmTMB(
  justif_violencia_protesta_consistente ~
    educ_cat_unordered * protesta_within +
    educ_cat_unordered * protesta_mean +
    edad + mujer + ideologia_std + factor(year) +
    (1 | idencuesta),
  data = elsoc_final_2,
  family = gaussian()
)

mod_rewb_int_estatal <- glmmTMB(
  justif_violencia_estatal ~
    educ_cat_unordered * protesta_within +
    educ_cat_unordered * protesta_mean +
    edad + mujer + ideologia_std + factor(year) +
    (1 | idencuesta),
  data = elsoc_final_2,
  family = gaussian()
)
```

```{r}
#| label: tbl-modelos-rewb
#| tbl-cap: "Within-between models: decomposing participation into within- and between-person components"
#| echo: false
#| warning: false
#| results: asis

ccoef_rewb <- list(
  "educ_cat_unorderedTechnical College incomplete" = "Technical College incomplete",
  "educ_cat_unorderedTechnical College complete"   = "Technical College complete",
  "educ_cat_unorderedUniversity incomplete"        = "University incomplete",
  "educ_cat_unorderedUniversity complete"          = "University complete",
  "protesta_within"                                = "Participation (within-person)",
  "protesta_mean"                                  = "Participation (between-person mean)",
  "edad"                                           = "Age",
  "mujer"                                          = "Woman",
  "ideologia_std"                                  = "Ideology",
  "factor(year)2017"                               = "2017",
  "factor(year)2018"                               = "2018",
  "factor(year)2019"                               = "2019",
  "factor(year)2022"                               = "2022",
  "factor(year)2023"                               = "2023"
)

{
  .tex <- capture.output(
    texreg(list(mod_rewb_protesta, mod_rewb_estatal),
       custom.model.names = c("Model 3a", "Model 3b"),
       stars = c(0.001, 0.01, 0.05),
       omit.coef = "(Intercept)",
       custom.coef.map = ccoef_rewb,
       digits = 3,
       groups = list(
         "Education (Ref. = High School or less)" = 1:4,
         "Participation (decomposed)"             = 5:6,
         "Controls"                               = 7:9,
         "Year (Ref. = 2016)"                     = 10:14
       ),
       custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05. Within-person = deviation from individual mean; Between-person = individual mean across waves.",
       leading.zero = FALSE,
       use.packages = FALSE,
       booktabs = TRUE,
       dcolumn = FALSE,
       scalebox = 0.8,
       include.bic = FALSE,
       include.loglik = FALSE,
       include.aic = TRUE,
       center = TRUE,
       custom.gof.names = c("AIC", "Num. obs.", "Num. groups: idencuesta", "Var: idencuesta"),
       reorder.gof = c(3, 4, 2, 1, 5),
       custom.gof.rows = list("BIC" = c(BIC(mod_rewb_protesta), BIC(mod_rewb_estatal))),
       caption = NULL,
       table.placement = "",
       label = "",
       float.pos = "!htbp")
  )
  cat(.tex[!grepl("\\begin{table",  .tex, fixed = TRUE) &
           !grepl("\\end{table}",   .tex, fixed = TRUE) &
           !grepl("\\begin{center}", .tex, fixed = TRUE) &
           !grepl("\\end{center}",  .tex, fixed = TRUE)],
      sep = "\n")
}
```

@tbl-modelos-rewb decomposes the participation effect into within- and between-person components. For protest violence (Model 3a), both components are statistically significant. The within-person coefficient (β = .252, *p* < .001) indicates that individuals show higher justification in waves where they participate relative to their own non-participating waves — consistent with a socialization process. The between-person coefficient is substantially larger (β = .686, *p* < .001), indicating that individuals who participate more frequently across waves already hold considerably higher baseline justifications — consistent with selection. Both mechanisms thus operate simultaneously for protest violence, though the selection component explains the larger share.

For state violence (Model 3b), the decomposition reveals a different pattern. The within-person coefficient (β = −.044, *p* = ns) is not significant, while the between-person coefficient is significant and substantially negative (β = −.432, *p* < .001). This suggests that the negative association between participation and state violence justification documented in Models 1–2 reflects predominantly pre-existing attitudinal differences rather than within-person change produced by participation itself: individuals who participate more frequently across waves already show lower state violence justification, consistent with a selection-dominant process. These results qualify the interpretation of H2b — the observed pattern appears driven primarily by the sorting of attitudinally distinct individuals into protest, rather than by experiential moral reframing generated by participation.

```{r}
#| label: tbl-modelos-rewb-int
#| tbl-cap: "Within-between models with education × participation interaction (Models 4a–4b): decomposing socialization and selection by educational level"
#| echo: false
#| warning: false
#| results: asis

ccoef_rewb_int <- list(
  "protesta_within"                                                = "Within-person (ref. = High School or less)",
  "educ_cat_unorderedTechnical College incomplete:protesta_within" = "W: x Technical College incomplete",
  "educ_cat_unorderedTechnical College complete:protesta_within"   = "W: x Technical College complete",
  "educ_cat_unorderedUniversity incomplete:protesta_within"        = "W: x University incomplete",
  "educ_cat_unorderedUniversity complete:protesta_within"          = "W: x University complete",
  "protesta_mean"                                                  = "Between-person (ref. = High School or less)",
  "educ_cat_unorderedTechnical College incomplete:protesta_mean"   = "B: x Technical College incomplete",
  "educ_cat_unorderedTechnical College complete:protesta_mean"     = "B: x Technical College complete",
  "educ_cat_unorderedUniversity incomplete:protesta_mean"          = "B: x University incomplete",
  "educ_cat_unorderedUniversity complete:protesta_mean"            = "B: x University complete"
)

{
  .tex <- capture.output(
    texreg(list(mod_rewb_int_protesta, mod_rewb_int_estatal),
       custom.model.names = c("Model 4a", "Model 4b"),
       stars = c(0.001, 0.01, 0.05),
       omit.coef = "(Intercept)",
       custom.coef.map = ccoef_rewb_int,
       digits = 3,
       groups = list(
         "Within-person (socialization)" = 1:5,
         "Between-person (selection)"    = 6:10
       ),
       custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05. Interaction coefficients represent differential effects relative to the High School reference. Education main effects, controls, and year dummies omitted.",
       leading.zero = FALSE,
       use.packages = FALSE,
       booktabs = TRUE,
       dcolumn = FALSE,
       scalebox = 0.85,
       include.bic = FALSE,
       include.loglik = FALSE,
       include.aic = TRUE,
       center = TRUE,
       custom.gof.names = c("AIC", "Num. obs.", "Num. groups: idencuesta", "Var: idencuesta"),
       reorder.gof = c(3, 4, 2, 1, 5),
       custom.gof.rows = list("BIC" = c(BIC(mod_rewb_int_protesta), BIC(mod_rewb_int_estatal))),
       caption = NULL,
       table.placement = "",
       label = "",
       float.pos = "!htbp")
  )
  cat(.tex[!grepl("\\begin{table",  .tex, fixed = TRUE) &
           !grepl("\\end{table}",   .tex, fixed = TRUE) &
           !grepl("\\begin{center}", .tex, fixed = TRUE) &
           !grepl("\\end{center}",  .tex, fixed = TRUE)],
      sep = "\n")
}
```

@tbl-modelos-rewb-int extends the decomposition by interacting education with both participation components, yielding a critical test of whether the educational paradox in protest violence reflects differential socialization or differential sorting. For protest violence (Model 4a), the evidence most consistently supports a sorting account: the education × within-person interaction for university complete is negligible and non-significant (β = .004, *p* = .945), while the education × between-person interaction is substantial and significant (β = .377, *p* < .001). University-educated protesters show essentially the same within-person socialization effect as lower-educated groups — participation shifts their attitudes no more than it does for the reference category — but they enter each wave with substantially higher baseline protest violence justification. The educational paradox documented in @tbl-modelos-interaccion is most consistent with a selection account: the cultural capital of university education appears to sort individuals into protest from a position of pre-existing permissiveness toward disruptive violence, rather than amplifying their responsiveness to participation.

For state violence (Model 4b), the interaction tells a more nuanced story. For lower-educated groups, the within-person effect of participation on state violence remains near zero and non-significant (base β = .039, ns). However, university-incomplete and university-complete groups show significant differential within-person effects (β = −.212, *p* = .016, and β = −.136, *p* = .032, respectively), implying total within-person effects of −.173 and −.097 — genuine attitudinal shifts when these individuals transition between participating and non-participating waves. This suggests that participation does produce some within-person reduction in state violence tolerance specifically among the most educated, qualifying the pure selection story. Nonetheless, the dominant mechanism for state violence remains sorting: the between-person effect is substantial for the reference category (β = −.368, *p* < .001) and further amplified for university completers (additional β = −.212, *p* = .059). The overall picture thus combines a selection-dominant pattern with a partial socialization process that operates exclusively at the upper end of the educational distribution.

These results also address a potential concern about the power of within-person estimates. Of the 3,627 individuals meeting the minimum three-wave inclusion criterion, 244 were subsequently excluded via listwise deletion on model covariates (ideology, age, gender, and outcome variables), yielding the analytical sample of 3,383 used in all regression models. Within the broader criterion-eligible panel of 3,627, 1,262 (34.8%) changed participation status at least once across waves, and within-person variation accounts for 54.8% of total participation variance. The within-null for state violence in Model 3b therefore cannot be attributed to insufficient variation in participation across waves.

```{r}
#| label: fig-efectos-marginales
#| fig-cap: "Marginal effects of protest participation on violence justification by education"
#| fig-width: 14
#| fig-height: 10
#| echo: false
#| warning: false

# ===================================================================
# PANEL A: Marginal effect of participation by EDUCATION
# for PROTEST Violence
# ===================================================================

# Calculate marginal effects manually
# Marginal effect = β_protest + β_interaction * education_level

# Extract model coefficients
coef_educ_prot <- fixef(mod_educ_protesta_int)$cond

# Base marginal effect (secondary education = reference category)
me_base <- coef_educ_prot["protesta_dummy"]

# Marginal effects for each education level
me_data_educ_prot <- data.frame(
  education = c("High School\nor less", "Tech. College\nincomplete", "Tech. College\ncomplete",
                "University\nincomplete", "University\ncomplete"),
  me = c(
    me_base,  # Secondary (reference)
    me_base + coef_educ_prot["educ_cat_unorderedTechnical College incomplete:protesta_dummy"],
    me_base + coef_educ_prot["educ_cat_unorderedTechnical College complete:protesta_dummy"],
    me_base + coef_educ_prot["educ_cat_unorderedUniversity incomplete:protesta_dummy"],
    me_base + coef_educ_prot["educ_cat_unorderedUniversity complete:protesta_dummy"]
  )
)

# Calculate approximate standard errors using vcov
vcov_mat <- vcov(mod_educ_protesta_int)$cond

# SE for reference category (variance of protesta_dummy only)
se_base <- sqrt(vcov_mat["protesta_dummy", "protesta_dummy"])

# For interactions: SE = sqrt(Var(β1) + Var(β2) + 2*Cov(β1,β2))
se_tec_inc <- sqrt(
  vcov_mat["protesta_dummy", "protesta_dummy"] +
  vcov_mat["educ_cat_unorderedTechnical College incomplete:protesta_dummy",
           "educ_cat_unorderedTechnical College incomplete:protesta_dummy"] +
  2 * vcov_mat["protesta_dummy", "educ_cat_unorderedTechnical College incomplete:protesta_dummy"]
)

se_tec_comp <- sqrt(
  vcov_mat["protesta_dummy", "protesta_dummy"] +
  vcov_mat["educ_cat_unorderedTechnical College complete:protesta_dummy",
           "educ_cat_unorderedTechnical College complete:protesta_dummy"] +
  2 * vcov_mat["protesta_dummy", "educ_cat_unorderedTechnical College complete:protesta_dummy"]
)

se_univ_inc <- sqrt(
  vcov_mat["protesta_dummy", "protesta_dummy"] +
  vcov_mat["educ_cat_unorderedUniversity incomplete:protesta_dummy",
           "educ_cat_unorderedUniversity incomplete:protesta_dummy"] +
  2 * vcov_mat["protesta_dummy", "educ_cat_unorderedUniversity incomplete:protesta_dummy"]
)

se_univ_comp <- sqrt(
  vcov_mat["protesta_dummy", "protesta_dummy"] +
  vcov_mat["educ_cat_unorderedUniversity complete:protesta_dummy",
           "educ_cat_unorderedUniversity complete:protesta_dummy"] +
  2 * vcov_mat["protesta_dummy", "educ_cat_unorderedUniversity complete:protesta_dummy"]
)

me_data_educ_prot$se <- c(se_base, se_tec_inc, se_tec_comp, se_univ_inc, se_univ_comp)
me_data_educ_prot$ci_low <- me_data_educ_prot$me - 1.96 * me_data_educ_prot$se
me_data_educ_prot$ci_high <- me_data_educ_prot$me + 1.96 * me_data_educ_prot$se
me_data_educ_prot$significant <- !(me_data_educ_prot$ci_low < 0 & me_data_educ_prot$ci_high > 0)

# Reorder levels
me_data_educ_prot$education <- factor(me_data_educ_prot$education,
                                       levels = c("High School\nor less", "Tech. College\nincomplete",
                                                  "Tech. College\ncomplete", "University\nincomplete",
                                                  "University\ncomplete"))

# Panel A plot
panel_a <- ggplot(me_data_educ_prot, aes(x = education, y = me)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 0.8) +
  geom_point(aes(color = significant), size = 3.5) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high, color = significant),
                width = 0.2, linewidth = 1) +
  scale_color_manual(values = c("FALSE" = "gray50", "TRUE" = "#2C7BB6"),
                     guide = "none") +
  labs(title = "Protest Violence",
       x = "Education Level",
       y = "Marginal Effect of Participation\n(change in justification)") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 11),
        plot.title = element_text(size = 12, face = "bold"),
        panel.grid.minor = element_blank())

# ===================================================================
# PANEL B: Marginal effect of participation by EDUCATION
# for STATE Violence
# ===================================================================

coef_educ_est <- fixef(mod_educ_estatal_int)$cond
me_base_est <- coef_educ_est["protesta_dummy"]

me_data_educ_est <- data.frame(
  education = c("High School\nor less", "Tech. College\nincomplete", "Tech. College\ncomplete",
                "University\nincomplete", "University\ncomplete"),
  me = c(
    me_base_est,
    me_base_est + coef_educ_est["educ_cat_unorderedTechnical College incomplete:protesta_dummy"],
    me_base_est + coef_educ_est["educ_cat_unorderedTechnical College complete:protesta_dummy"],
    me_base_est + coef_educ_est["educ_cat_unorderedUniversity incomplete:protesta_dummy"],
    me_base_est + coef_educ_est["educ_cat_unorderedUniversity complete:protesta_dummy"]
  )
)

vcov_mat_est <- vcov(mod_educ_estatal_int)$cond
se_base_est <- sqrt(vcov_mat_est["protesta_dummy", "protesta_dummy"])

se_tec_inc_est <- sqrt(
  vcov_mat_est["protesta_dummy", "protesta_dummy"] +
  vcov_mat_est["educ_cat_unorderedTechnical College incomplete:protesta_dummy",
               "educ_cat_unorderedTechnical College incomplete:protesta_dummy"] +
  2 * vcov_mat_est["protesta_dummy", "educ_cat_unorderedTechnical College incomplete:protesta_dummy"]
)

se_tec_comp_est <- sqrt(
  vcov_mat_est["protesta_dummy", "protesta_dummy"] +
  vcov_mat_est["educ_cat_unorderedTechnical College complete:protesta_dummy",
               "educ_cat_unorderedTechnical College complete:protesta_dummy"] +
  2 * vcov_mat_est["protesta_dummy", "educ_cat_unorderedTechnical College complete:protesta_dummy"]
)

se_univ_inc_est <- sqrt(
  vcov_mat_est["protesta_dummy", "protesta_dummy"] +
  vcov_mat_est["educ_cat_unorderedUniversity incomplete:protesta_dummy",
               "educ_cat_unorderedUniversity incomplete:protesta_dummy"] +
  2 * vcov_mat_est["protesta_dummy", "educ_cat_unorderedUniversity incomplete:protesta_dummy"]
)

se_univ_comp_est <- sqrt(
  vcov_mat_est["protesta_dummy", "protesta_dummy"] +
  vcov_mat_est["educ_cat_unorderedUniversity complete:protesta_dummy",
               "educ_cat_unorderedUniversity complete:protesta_dummy"] +
  2 * vcov_mat_est["protesta_dummy", "educ_cat_unorderedUniversity complete:protesta_dummy"]
)

me_data_educ_est$se <- c(se_base_est, se_tec_inc_est, se_tec_comp_est, se_univ_inc_est, se_univ_comp_est)
me_data_educ_est$ci_low <- me_data_educ_est$me - 1.96 * me_data_educ_est$se
me_data_educ_est$ci_high <- me_data_educ_est$me + 1.96 * me_data_educ_est$se
me_data_educ_est$significant <- !(me_data_educ_est$ci_low < 0 & me_data_educ_est$ci_high > 0)

me_data_educ_est$education <- factor(me_data_educ_est$education,
                                      levels = c("High School\nor less", "Tech. College\nincomplete",
                                                 "Tech. College\ncomplete", "University\nincomplete",
                                                 "University\ncomplete"))

panel_b <- ggplot(me_data_educ_est, aes(x = education, y = me)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 0.8) +
  geom_point(aes(color = significant), size = 3.5) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high, color = significant),
                width = 0.2, linewidth = 1) +
  scale_color_manual(values = c("FALSE" = "gray50", "TRUE" = "#2C7BB6"),
                     guide = "none") +
  labs(title = "State Violence",
       x = "Education Level",
       y = "Marginal Effect of Participation\n(change in justification)") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 11),
        plot.title = element_text(size = 12, face = "bold"),
        panel.grid.minor = element_blank())

# Combine panels
combined_marginal_effects <- (panel_a | panel_b) +
  plot_annotation(
    tag_levels = 'A',
    caption = "Marginal effects with 95% CI. Red line indicates null effect. Grey points: non-significant.\nMarginal effect = change in justification from participating vs. not participating, conditioned by education. Compiled by the author based on ELSOC.",
    theme = theme(plot.caption = element_text(size = 9, hjust = 0))
  )

combined_marginal_effects
```

@fig-efectos-marginales visualizes the marginal effects of participation by education, allowing direct evaluation of the hypotheses about differential effects.

In Panel A, all education levels show significant positive associations between participation and protest violence justification, with magnitudes ranging from +0.27 (secondary education) to +0.55 (technical college incomplete). The association is stronger among technical college (+0.55) and university (+0.40) than among secondary education (+0.27), indicating that education amplifies the effect of participation rather than dampening it — a pattern at odds with the civilizing effect hypothesis among those who mobilize.

Panel B shows that only university-educated individuals have significant negative marginal effects on state violence justification (−0.20 to −0.22), while secondary and technical education show null effects (grey points on the red line). University-educated participants thus show higher protest violence justification (+0.40) while simultaneously showing lower state violence justification (−0.22). As the within-between analysis in Section 5.2.1 suggests, this state violence pattern is most consistent with a sorting account — individuals who tend to participate more appear to already hold lower state violence justification — rather than attitudinal change produced by participation itself.


### The Education "Paradox Effect"

```{r}
#| label: fig-paradoja-educacion
#| fig-cap: "Predicted values of violence justification by education and participation type"
#| fig-width: 12
#| fig-height: 6
#| echo: false
#| warning: false

# Generate predictions from the education model with interactions
pred_df <- ggpredict(mod_educ_protesta_int,
                     terms = c("educ_cat_unordered", "protesta_dummy")) %>%
  as.data.frame() %>%
  rename(participation = group) %>%
  mutate(participation = factor(participation,
                                 levels = c("0", "1"),
                                 labels = c("Did not participate", "Participated")))

# Panel A: Predictions
panel_a <- ggplot(pred_df, aes(x = x, y = predicted,
                                color = participation, group = participation,
                                fill = participation)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              alpha = 0.2, color = NA) +
  geom_line(aes(linetype = participation), linewidth = 1) +
  geom_point(aes(shape = participation), size = 2.5) +
  scale_color_manual(values = c("Did not participate" = "#F8766D", "Participated" = "#00BFC4"),
                     name = "Participation") +
  scale_fill_manual(values = c("Did not participate" = "#F8766D", "Participated" = "#00BFC4"),
                    name = "Participation") +
  scale_linetype_manual(values = c("Did not participate" = "dashed", "Participated" = "solid"),
                        name = "Participation") +
  scale_shape_manual(values = c("Did not participate" = 17, "Participated" = 16),
                     name = "Participation") +
  labs(
    title = "Protest Violence",
    x = "Education Level",
    y = "Justification (predicted)"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    legend.position = "bottom",
    legend.text = element_text(size = 11),
    plot.title = element_text(size = 13, face = "bold")
  )

# Panel B: Predicted values for state violence
pred_df_est <- ggpredict(mod_educ_estatal_int,
                         terms = c("educ_cat_unordered", "protesta_dummy")) %>%
  as.data.frame() %>%
  rename(participation = group) %>%
  mutate(participation = factor(participation,
                                 levels = c("0", "1"),
                                 labels = c("Did not participate", "Participated")))

panel_b <- ggplot(pred_df_est, aes(x = x, y = predicted,
                                    color = participation, group = participation,
                                    fill = participation)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              alpha = 0.2, color = NA) +
  geom_line(aes(linetype = participation), linewidth = 1) +
  geom_point(aes(shape = participation), size = 2.5) +
  scale_color_manual(values = c("Did not participate" = "#F8766D", "Participated" = "#00BFC4"),
                     name = "Participation") +
  scale_fill_manual(values = c("Did not participate" = "#F8766D", "Participated" = "#00BFC4"),
                    name = "Participation") +
  scale_linetype_manual(values = c("Did not participate" = "dashed", "Participated" = "solid"),
                        name = "Participation") +
  scale_shape_manual(values = c("Did not participate" = 17, "Participated" = 16),
                     name = "Participation") +
  labs(
    title = "State Violence",
    x = "Education Level",
    y = "Justification (predicted)"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    legend.position = "bottom",
    legend.text = element_text(size = 11),
    plot.title = element_text(size = 13, face = "bold")
  )

# Combine side by side with shared legend
plot_final <- (panel_a | panel_b) +
  plot_layout(guides = "collect") +
  plot_annotation(
    tag_levels = 'A',
    caption = "Predicted values with 95% CI. Shaded areas: confidence intervals. Other predictors at mean values. Compiled by the author based on ELSOC 2016–2023.",
    theme = theme(plot.caption = element_text(size = 7, hjust = 0))
  ) &
  theme(legend.position = "bottom")

# Display plot
plot_final
```

@fig-paradoja-educacion shows both the "educational paradox" (H1a–H1b) and the bidirectional reconfiguration (H2a–H2b) through direct comparison of violence types. Panel A shows the pattern for protest violence: among non-participants, higher education is associated with lower justification (descending red line from ~1.95 at secondary education to ~1.75 at university complete), consistent with the civilizing effect. Among participants, however, the direction reverses (ascending blue line from ~2.20 at secondary education to ~2.35 at technical college incomplete), suggesting that mobilization experience is associated with higher justification even among the most educated groups.

Panel B reveals the contrasting pattern for state violence, consistent with H2a–H2b. Among non-participants, both lines remain nearly parallel across education levels with no meaningful gradient, indicating that education alone does not generate a critical framework toward repression. Among participants, the blue line declines markedly at university levels — the only educational group for which participation is associated with lower state violence justification (university complete: −0.18***; university incomplete: −0.16*). Read together, the two panels are consistent with the bidirectionality of the reconfiguration: university-educated participants simultaneously show higher protest violence justification and lower state violence justification, consistent with the dual moral framework predicted in H2a–H2b.

## Bidirectional Reconfiguration of Violence Justification (Protest Violence vs. State Violence)

The interaction models (Model 1b and Model 2b) reveal the bidirectional dynamic predicted in H2a–H2b (university-educated). H2a is supported: for protest violence, university education is associated with lower justification among non-participants ($-0.13***$), but participation is associated with an increase ($+0.27***$) that is larger at higher education levels (technical college incomplete: $+0.28*$; university complete: $+0.13**$), consistent with the nullification of the civilizing effect. H2b is supported in terms of the observed association: for state violence, education alone ($+0.06$, ns) and participation alone ($-0.04$, ns) show no independent effects, but their negative interactions produce significant lower justification among university-educated individuals (incomplete: $-0.16*$; complete: $-0.18***$), indicating that the combination of university education and participation is associated with lower acceptance of state repression. As shown in the within-between analysis (Section 5.2.1), however, the within-person component of this state violence effect is non-significant, suggesting that this pattern predominantly reflects pre-existing attitudinal differences among habitual protesters rather than a socialization process generated by participation itself.

The controls are consistent with the impact of the 2019 outbreak: moderate increase in protest justification ($+0.05*$) but sharp fall in state violence ($-0.40***$), substantially reconfiguring attitudes toward police repression. Ideology operates in opposite directions: right-wing ideology reduces justification of protests ($-0.03***$) but increases that of state violence ($+0.05***$). Women reject state violence more ($-0.13***$) with no differences in protests. This pattern indicates that the bidirectional reconfiguration predicted in H2a–H2b is most pronounced among those who combine university education with direct experience of mobilization. The protest violence component of this reconfiguration involves both socialization and selection, while the state violence component appears driven primarily by selection — a distinction documented in the within-between models.

```{r}
#| label: fig-bidireccional-tiempo
#| fig-cap: "Temporal evolution of bidirectional reconfiguration: secondary vs. university education (2016–2023)"
#| fig-width: 12
#| fig-height: 9
#| dpi: 300
#| echo: false
#| warning: false

# Get predictions for protest violence
pred_protestas_tiempo <- ggpredict(mod_educ_protesta_int,
                                   terms = c("year", "protesta_dummy", "educ_cat_unordered [High School or less, University complete]"),
                                   typical = "mean")

# Get predictions for state violence
pred_estatal_tiempo <- ggpredict(mod_educ_estatal_int,
                                 terms = c("year", "protesta_dummy", "educ_cat_unordered [High School or less, University complete]"),
                                 typical = "mean")

# Convert to dataframes
pred_protestas_df <- as.data.frame(pred_protestas_tiempo)
pred_protestas_df$violence_type <- "Protest Violence"
pred_protestas_df$participation <- factor(pred_protestas_df$group,
                                          labels = c("Did not participate", "Participated"))
pred_protestas_df$education <- pred_protestas_df$facet

pred_estatal_df <- as.data.frame(pred_estatal_tiempo)
pred_estatal_df$violence_type <- "State Violence"
pred_estatal_df$participation <- factor(pred_estatal_df$group,
                                        labels = c("Did not participate", "Participated"))
pred_estatal_df$education <- pred_estatal_df$facet

# Combine both datasets
pred_combined <- rbind(pred_protestas_df, pred_estatal_df)

# Create plot with facets by education and participation
ggplot(pred_combined, aes(x = x, y = predicted, color = violence_type, group = violence_type)) +
  geom_line(aes(linetype = violence_type), linewidth = 1.2) +
  geom_point(aes(shape = violence_type), size = 2.5) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = violence_type),
              alpha = 0.15, color = NA) +
  facet_grid(participation ~ education,
             labeller = labeller(
               participation = c("Did not participate" = "Non-Participants",
                                "Participated" = "Participants")
             )) +
  scale_color_manual(values = c("Protest Violence" = "#EFC000FF",
                                 "State Violence" = "#0073C2FF"),
                     name = "Type of Violence") +
  scale_fill_manual(values = c("Protest Violence" = "#EFC000FF",
                                "State Violence" = "#0073C2FF"),
                    name = "Type of Violence") +
  scale_linetype_manual(values = c("Protest Violence" = "solid",
                                    "State Violence" = "dashed"),
                        name = "Type of Violence") +
  scale_shape_manual(values = c("Protest Violence" = 16,
                                 "State Violence" = 17),
                     name = "Type of Violence") +
  geom_vline(xintercept = 2019, linetype = "dashed", color = "red", alpha = 0.5) +
  labs(
    title = "Bidirectional Reconfiguration: Secondary vs. University Education (2016–2023)",
    x = "Year",
    y = "Predicted Violence Justification",
    caption = "Dashed red line: 2019 (social outbreak). Shaded areas: 95% CI. Other predictors at mean values.\nCompiled by the author based on ELSOC 2016–2023."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 10),
    strip.text = element_text(face = "bold", size = 11),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    plot.caption = element_text(size = 8)
  ) +
  ylim(1.3, 2.4)
```

@fig-bidireccional-tiempo visualizes the temporal dynamics of the bidirectional reconfiguration predicted in H2a–H2b through the direct contrast between secondary and university education — the two polar cases that most clearly reveal whether the dual moral framework requires university socialization.

Among non-participants (top row), both education levels show parallel lines between protest and state violence throughout the entire period, confirming that participation is a necessary condition for the bidirectional reconfiguration: without mobilization experience, neither educational level generates the divergence between violence types predicted in H2a–H2b.

Among participants (bottom row), the contrast is clear. Secondary-educated participants show elevated protest violence justification post-2019 (consistent with H2a) but near-parallel lines with state violence, which does not support H2b: they show higher acceptance of tactical violence but without a corresponding rejection of state repression. University-educated participants display the pattern consistent with the full dual framework: (1) a sustained elevated yellow line (~2.0–2.1), at odds with the civilizing effect and consistent with H2a; (2) a notable decline of the blue line in 2019 from ~1.75 to ~1.35 ($-0.40***$), consistent with H2b; (3) a post-2019 divergence of ~0.7 points between both violence types. This direct comparison is consistent with H2a and H2b being jointly observable only under the combination of university education and mobilization experience.
