```{r}
#| include: false
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

library(dplyr)
library(ggplot2)
library(glmmTMB)
library(texreg)
library(ggeffects)
library(patchwork)
library(gt)
library(kableExtra)
library(here)
library(margins)
library(broom)
library(modelsummary)
library(stargazer)
library(tidyr)

# Load data
load(here("input/data/proc/elsoc_final_2.RData"))
```

# Results

This section presents the findings on the factors shaping the justification of violence in the context of social protest in Chile.

```{r}
#| results: false


# 1st Step: Convert to non-ordered factor
elsoc_final_2 <- elsoc_final_2 %>%
  mutate(
    educ_cat_unordered = factor(educ_cat_factor, ordered = FALSE),
    movilidad_cat_unordered = factor(movilidad_cat_factor, ordered = FALSE)
  )

# 2nd Step: Create consistent index of violence in protests (only variables in all waves)
elsoc_final_2 <- elsoc_final_2 %>%
  mutate(
    # Consistent temporal index (students and workers only)
    justif_violencia_protesta_consistente = rowMeans(
      dplyr::select(., violencia_trabajadores, violencia_estudiantes),
      na.rm = TRUE
    ),
    # Broad index (all variables available per wave)
    justif_violencia_protesta_amplio = justif_violencia_protesta
  )

# Check availability per wave
elsoc_final_2 %>%
  group_by(year) %>%
  summarise(
    n = n(),
    estudiantes = sum(!is.na(violencia_estudiantes)),
    trabajadores = sum(!is.na(violencia_trabajadores)),
    inmobiliario = sum(!is.na(violencia_inmobiliario)),
    transporte = sum(!is.na(violencia_transporte)),
    locales = sum(!is.na(violencia_locales)),
    carab_marchas = sum(!is.na(violencia_carabineros_marchas)),
    carab_tomas = sum(!is.na(violencia_carabineros_tomas))
  ) %>%
  print()

```

## Descriptive Statistics

### Justification of Violence by Education Level

The patterns observed in @tbl-justificacion-educ reveal a notable trend. Among non-participants, higher education levels tend to be associated with lower justification of protest violence, with means declining from approximately 1.90 to 1.80. However, this trend reverses among those who do participate: participants with university education display the highest justification levels, reaching means around 2.29 and 2.27. This differentiated effect of education appears to be specific to protest violence, as, by contrast, the justification of state violence shows a much more stable pattern, with considerably smaller variation across educational levels.

```{r}
#| label: tbl-justificacion-educ
#| tbl-cap: "Justification of violence by education level and protest participation"
#| echo: false
#| warning: false
#| results: asis

# Redefine factor labels:
elsoc_final_2 <- elsoc_final_2 %>%
  mutate(
    educ_cat_unordered = factor(
      educ_cat_unordered,
      labels = c(
        "High School or less",
        "Technical College incomplete",
        "Technical College complete",
        "University incomplete",
        "University complete"
      )
    )
  )

# Create and display table
tabla_1 <- elsoc_final_2 %>%
  filter(!is.na(educ_cat_unordered) & !is.na(protesta_dummy)) %>%
  group_by(educ_cat_unordered, Participation = factor(protesta_dummy,
                                                       labels = c("Yes", "No"))) %>%
  summarise(
    N = n(),
    `Protest Violence` = round(mean(justif_violencia_protesta_consistente, na.rm = TRUE), 2),
    `SD` = round(sd(justif_violencia_protesta_consistente, na.rm = TRUE), 2),
    `State Violence` = round(mean(justif_violencia_estatal, na.rm = TRUE), 2),
    `SD ` = round(sd(justif_violencia_estatal, na.rm = TRUE), 2),
    .groups = "drop"
  ) %>%
  knitr::kable(
    col.names = c("Education", "Participation", "N", "Mean", "SD", "Mean", "SD"),
    align = c("l", "l", "c", "c", "c", "c", "c"),
    format = "latex",
    booktabs = TRUE
  ) %>%
  kable_styling(
    latex_options = c("striped", "hold_position", "scale_down"),
    font_size = 9,
    position = "center"
  ) %>%
  add_header_above(c(" " = 3, "Protest Violence" = 2, "State Violence" = 2)) %>%
  footnote(
    general = "Compiled by the author based on ELSOC (2016–2023). Scale of 1–5, where 1 = Never justified and 5 = Always justified",
    threeparttable = TRUE
  )

tabla_1
```


### Temporal Evolution: Justification of Violence and Protest Participation (2016–2023)

```{r}
#| label: fig-evolucion-temporal
#| fig-cap: "Evolution of protest participation 2016–2023"
#| fig-width: 12
#| fig-height: 7
#| echo: false

# Participation Panel
plot_evol_particip <- elsoc_final_2 %>%
  group_by(year) %>%
  summarise(
    Participation = 100 * mean(protesta_dummy, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = year, y = Participation)) +
  geom_vline(xintercept = 2019, linetype = "dashed", color = "red", alpha = 0.6) +
  geom_line(linewidth = 1.2, color = "#619CFF") +
  geom_point(size = 2.5, color = "#619CFF") +
  labs(
    title = "Protest Participation Over Time",
    x = "Year",
    y = "% who Participated") +
  theme_minimal(base_size = 14)

# Combine
grafico_evolucion <- plot_evol_particip +
  plot_annotation(
    tag_levels = 'A',
    caption = "Compiled by the author based on ELSOC 2016-2023. Note: 2019 = Social Outbreak",
    theme = theme(plot.caption = element_text(size = 11))
  )

# Display plot
grafico_evolucion
```

The temporal pattern analysis of @fig-evolucion-temporal shows a result consistent with the observed protest cycle context. In 2019 — the year of the social outbreak — the highest average protest participation rate is recorded, reflecting the intensity of mobilization. In the post-2019 period, although participation declines from its peak, it remains at slightly higher levels than those observed before 2019, until its fall between 2021 and 2023, where it reaches its lowest average across the entire reported cycle.

```{r}
#| label: fig-participacion-estratificada
#| fig-cap: "Sociodemographic composition of protest participants by year: structural changes post-outbreak. Compiled by the author based on ELSOC."
#| fig-width: 12
#| fig-height: 18
#| echo: false
#| warning: false

# Panel: Direct comparison Protest vs State violence by year
comparacion_data <- elsoc_final_2 %>%
  group_by(year) %>%
  summarise(
    `Protest Violence` = mean(justif_violencia_protesta_consistente, na.rm = TRUE),
    `State Violence` = mean(justif_violencia_estatal, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  tidyr::pivot_longer(cols = c("Protest Violence", "State Violence"),
                      names_to = "Type",
                      values_to = "Justification")

just_viol <- ggplot(comparacion_data, aes(x = year, y = Justification, color = Type, group = Type)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 3.5) +
  geom_vline(xintercept = 2019, linetype = "dashed", color = "red", alpha = 0.6) +
  scale_color_manual(values = c("Protest Violence" = "#EFC000FF",
                                  "State Violence" = "#0073C2FF"),
                     name = "Type of Violence") +
  labs(
    title = "Divergence between Justification of Violence in Protests and State Violence",
    x = "Year",
    y = "Justification (scale 1-5)",
    caption = "Red line: 2019 social outbreak. Compiled by the author based on ELSOC 2016–2023."
  ) +
  theme_minimal(base_size = 15) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 11),
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 11, color = "gray30"),
    plot.caption = element_text(size = 10, hjust = 0)
  ) +
  ylim(1.4, 2.1)

# Display plot
just_viol
```

@fig-participacion-estratificada reveals the radical transformation of moral hierarchies around political violence. Pre-2019 (2016–2018), both forms of violence were evaluated with similar criteria (gap of ±0.05 points), reflecting liberal-democratic consensus where violence "from below" and "from above" were equally rejected (~1.6–1.7). In 2019, the gap reverses dramatically: state violence was justified ~0.15 points more than protest violence, reflecting the initial shock and polarization in response to perceived "chaos."

However, in the post-2019 period (2022–2023), the lines cross definitively: protest violence is justified ~0.10–0.15 points more than state violence, and this gap stabilizes. This is the critical finding: the outbreak generated a permanent reconfiguration of collective moral hierarchies, in which disruptive tactical violence is normalized more than state repression. This pattern evidences the bidirectional reconfiguration: the emerging dual moral framework post-outbreak redefines the boundaries of legitimate political action depending on who exercises violence and in what context.

```{r}
#| include: false

# Function to calculate means and CIs by group and year
calc_means_ci <- function(data, group_var, tipo_violencia) {
  data %>%
    filter(!is.na(year) & year != 2019 & !is.na(!!sym(group_var)) & !is.na(protesta_dummy)) %>%
    group_by(year, !!sym(group_var), protesta_dummy) %>%
    summarise(
      mean = mean(!!sym(tipo_violencia), na.rm = TRUE),
      se = sd(!!sym(tipo_violencia), na.rm = TRUE) / sqrt(n()),
      n = n(),
      .groups = "drop"
    ) %>%
    mutate(
      ci_low = mean - 1.96 * se,
      ci_high = mean + 1.96 * se,
      participation = factor(protesta_dummy,
                            levels = c(0, 1),
                            labels = c("Did not participate", "Participated"))
    )
}

# Panel Education × Protest Violence
data_educ_prot <- calc_means_ci(elsoc_final_2, "educ_cat_unordered", "justif_violencia_protesta_consistente")

p1 <- ggplot(data_educ_prot, aes(x = educ_cat_unordered, y = mean, color = participation, group = participation)) +
  geom_point(position = position_dodge(width = 0.5), size = 2.5) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high),
                position = position_dodge(width = 0.5), width = 0.2, linewidth = 0.8) +
  geom_line(position = position_dodge(width = 0.5), linewidth = 1) +
  facet_wrap(~year, ncol = 3) +
  scale_color_manual(values = c("Did not participate" = "#F8766D", "Participated" = "#00BFC4")) +
  labs(title = "Education × Protest Violence",
       x = "Education Level",
       y = "Justification (scale 1-5)",
       color = "") +
  theme_minimal(base_size = 15) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        legend.position = "bottom",
        legend.text = element_text(size = 11),
        strip.text = element_text(face = "bold", size = 12)) +
  ylim(1, 3)

# Panel Education × State Violence
data_educ_est <- calc_means_ci(elsoc_final_2, "educ_cat_unordered", "justif_violencia_estatal")

p2 <- ggplot(data_educ_est, aes(x = educ_cat_unordered, y = mean, color = participation, group = participation)) +
  geom_point(position = position_dodge(width = 0.5), size = 2.5) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high),
                position = position_dodge(width = 0.5), width = 0.2, linewidth = 0.8) +
  geom_line(position = position_dodge(width = 0.5), linewidth = 1) +
  facet_wrap(~year, ncol = 3) +
  scale_color_manual(values = c("Did not participate" = "#F8766D", "Participated" = "#00BFC4")) +
  labs(title = "Education × State Violence",
       x = "Education Level",
       y = "Justification (scale 1-5)",
       color = "") +
  theme_minimal(base_size = 15) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        legend.position = "bottom",
        legend.text = element_text(size = 11),
        strip.text = element_text(face = "bold", size = 12)) +
  ylim(1, 3)

# Combine the 2 panels by education
educ_plot <- (p1 / p2) +
  plot_layout(guides = "collect") +
  plot_annotation(
    tag_levels = 'A',
    caption = "Points show means; bars show 95% CI. Year 2019 (social outbreak) excluded. Compiled by the author based on ELSOC."
  ) &
  theme(
    legend.position = "bottom",
    plot.caption = element_text(size = 7)
  )
```


```{r}
#| label: fig-estallido-1
#| fig-cap: "Comparison of violence justification before (2016–2018) and after (2022–2023) the social outbreak by education and participation"
#| fig-width: 12
#| fig-height: 20


# Display plot
educ_plot
```

@fig-estallido-1 presents the temporal evolution of violence justification by education, comparing year-by-year levels before (2016–2018) and after (2022–2023) the October 2019 social outbreak, revealing critical patterns about the persistence of the political shock effects.

Among non-participants, relative stability is observed in the pre-2019 period, with justification levels oscillating between ~1.5–1.8 for higher education and ~1.7–2.0 for lower education. In the post-2019 period (2022–2023), all educational groups show a moderate increase: university completers rise from ~1.5 to ~1.8, while basic/secondary education remains in the ~1.8–2.0 range. Among participants, the pattern shows sustained convergence: from 2016 to 2023, all educational levels maintain moderately high justifications (~2.0–2.3) with overlapping confidence intervals, indicating that the experience of mobilization equalizes attitudes regardless of cultural capital. Crucially, the gaps between participants and non-participants remain consistent year after year in the post-2019 period, showing that the participation effect persists stably years after the outbreak.

The contrast reveals that among non-participants, a dramatic jump is observed between 2018 and 2022 across all educational groups: university completers go from ~1.3–1.5 (2016–2018) to ~1.9–2.0 (2022–2023). This is the most critical finding: the outbreak did not generate permanent rejection of state repression, but rather normalized its justification even among privileged sectors, and this normalization remains stable in 2022 and 2023. Among participants, the pattern reverses completely: all educational groups show sustained declines post-2019, going from ~1.8–2.0 (2016–2018) to ~1.5–1.7 (2022–2023), with university completers maintaining consistently low levels (~1.4). The year-by-year trajectory confirms that direct experience of repression during the outbreak generated a permanent "experiential delegitimization" effect among demonstrators, consistently observable in both post-outbreak years.

The data reveal that the 2019 outbreak was not a temporary shock but a permanent inflection point that structurally reconfigured dual moral frameworks. The year-by-year visualization demonstrates that the divergence between participants and non-participants is not transient fluctuation but stable polarization: while participants develop lasting rejection of state violence yet maintain justification of protest violence, non-participants experience divergent patterns depending on cultural capital. These frameworks remain stable 3–4 years after the critical event, crystallizing the hypothesis of permanent bidirectional polarization.


## Exploratory Statistics

```{r}
#| echo: false
#| include: false

# ============================================
# EDUCATION MODELS
# ============================================

# Model 1a: Education - Main effects for Protest Violence
mod_educ_protesta_main <- glmmTMB(
  justif_violencia_protesta_consistente ~ educ_cat_unordered + protesta_dummy +
    edad + mujer + ideologia_std + factor(year) +
    (1 | idencuesta),
  data = elsoc_final_2,
  family = gaussian()
)

# Model 1b: Education × Protest for Protest Violence (with interaction)
mod_educ_protesta_int <- glmmTMB(
  justif_violencia_protesta_consistente ~ educ_cat_unordered * protesta_dummy +
    edad + mujer + ideologia_std + factor(year) +
    (1 | idencuesta),
  data = elsoc_final_2,
  family = gaussian()
)

# Model 2a: Education - Main effects for State Violence
mod_educ_estatal_main <- glmmTMB(
  justif_violencia_estatal ~ educ_cat_unordered + protesta_dummy +
    edad + mujer + ideologia_std + factor(year) +
    (1 | idencuesta),
  data = elsoc_final_2,
  family = gaussian()
)

# Model 2b: Education × Protest for State Violence (with interaction)
mod_educ_estatal_int <- glmmTMB(
  justif_violencia_estatal ~ educ_cat_unordered * protesta_dummy +
    edad + mujer + ideologia_std + factor(year) +
    (1 | idencuesta),
  data = elsoc_final_2,
  family = gaussian()
)
```

```{r}
#| label: tbl-modelos-principales
#| tbl-cap: "Main models of violence justification"
#| echo: false
#| warning: false
#| results: asis

# Custom coefficient map
ccoef <- list(
  "educ_cat_unorderedTechnical College incomplete" = "Technical College incomplete",
  "educ_cat_unorderedTechnical College complete" = "Technical College complete",
  "educ_cat_unorderedUniversity incomplete" = "University incomplete",
  "educ_cat_unorderedUniversity complete" = "University complete",
  "protesta_dummy" = "Protest participation",
  "edad" = "Age",
  "mujer" = "Woman",
  "ideologia_std" = "Ideology",
  "factor(year)2017" = "2017",
  "factor(year)2018" = "2018",
  "factor(year)2019" = "2019",
  "factor(year)2022" = "2022",
  "factor(year)2023" = "2023"
)

texreg(list(mod_educ_protesta_main, mod_educ_estatal_main),
    custom.model.names = c("Model 1", "Model 2"),
       stars = c(0.001, 0.01, 0.05),
       omit.coef = "(Intercept)",
       custom.coef.map = ccoef,
       digits = 3,
       groups = list("Education (Ref. = High School or less)" = 1:4,
         "Controls" = 5:8,
         "Year (Ref. = 2016)" = 9:13),
       custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05",
       leading.zero = FALSE,
       use.packages = FALSE,
       booktabs = TRUE,
       dcolumn = FALSE,
      scalebox = 0.8,
       include.bic = TRUE,
       include.loglik = FALSE,
       include.aic = TRUE,
       center = TRUE,
       custom.gof.names = c("AIC", "BIC", "Num. obs.", "Num. groups: idencuesta"),
       reorder.gof = c(3, 4, 1, 2),
       table.placement = "",
       label = "",
       float.pos = "!htbp")
```


@tbl-modelos-principales presents two main models guiding the analysis of this study. Model 1 examines the effects of education and protest participation on protest violence justification, and Model 2 evaluates those same effects on state violence justification. Both models include random effects by individual and controls for age, gender, ideology, and year.

The results reveal differentiated patterns depending on the type of violence. For protest violence, participation shows a consistent positive effect, increasing justification by approximately 0.36 points. By contrast, for state violence, the direct effect of participation is negative and highly significant, showing that protest participants do not more readily justify police repression.

The controls operate consistently: right-wing ideology reduces justification of protest violence (~-0.03 per point on the scale) but increases it for state violence (~+0.05). Women reject state violence more (~-0.13) with no differences on protest violence. The year 2019 shows the most dramatic contrast: a moderate increase in protest violence justification but a sharp fall in state violence justification (~-0.40), reflecting the impact of the social outbreak on attitudes toward police repression.

```{r}
#| label: tbl-modelos-interaccion
#| tbl-cap: "Interaction models for violence justification"
#| echo: false
#| warning: false
#| results: asis

# Custom coefficient map for interaction models
ccoef_int <- list(
  "educ_cat_unorderedTechnical College incomplete" = "Technical College incomplete",
  "educ_cat_unorderedTechnical College complete" = "Technical College complete",
  "educ_cat_unorderedUniversity incomplete" = "University incomplete",
  "educ_cat_unorderedUniversity complete" = "University complete",
  "protesta_dummy" = "Protest participation",
  "educ_cat_unorderedTechnical College incomplete:protesta_dummy" = "Technical College incomplete × Protest",
  "educ_cat_unorderedTechnical College complete:protesta_dummy" = "Technical College complete × Protest",
  "educ_cat_unorderedUniversity incomplete:protesta_dummy" = "University incomplete × Protest",
  "educ_cat_unorderedUniversity complete:protesta_dummy" = "University complete × Protest",
  "edad" = "Age",
  "mujer" = "Woman",
  "ideologia_std" = "Ideology",
  "factor(year)2017" = "2017",
  "factor(year)2018" = "2018",
  "factor(year)2019" = "2019",
  "factor(year)2022" = "2022",
  "factor(year)2023" = "2023"
)

texreg(list(mod_educ_protesta_int, mod_educ_estatal_int),
    custom.model.names = c("Model 1b", "Model 2b"),
       stars = c(0.001, 0.01, 0.05),
       omit.coef = "(Intercept)",
       custom.coef.map = ccoef_int,
       digits = 3,
       groups = list("Education (Ref. = High School or less)" = 1:4,
         "Education Interactions" = 6:9,
         "Controls" = 10:12,
         "Year (Ref. = 2016)" = 13:17),
       custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05",
       leading.zero = FALSE,
       use.packages = FALSE,
       booktabs = TRUE,
       dcolumn = FALSE,
      scalebox = 0.8,
       include.bic = TRUE,
       include.loglik = FALSE,
       include.aic = TRUE,
       center = TRUE,
       custom.gof.names = c("AIC", "BIC", "Num. obs.", "Num. groups: idencuesta"),
       reorder.gof = c(3, 4, 1, 2),
       table.placement = "",
       label = "",
       float.pos = "!htbp")
```

The models in @tbl-modelos-interaccion reveal that the effect of participation varies significantly by education level. For protest violence, the interactions between education and participation are positive, indicating that the legitimizing effect of participating is stronger among more educated individuals. A person with complete university education who participates increases their justification by 0.40 points versus 0.27 for secondary education, evidencing the "educational paradox": education reduces justification among non-participants (-0.13***) but amplifies the effect of participation.

For state violence, the pattern reverses completely. The interactions are negative (university incomplete: -0.16*; university complete: -0.18***), indicating that only the combination of university education + participation generates significant rejection of repression (total effect: -0.22 points). Education alone (+0.06, ns) or participation alone (-0.04, ns) are insufficient; it is their interaction that activates the differentiated rejection.

The bidirectional reconfiguration (legitimizing protests, rejecting repression) is clearest among university-educated participants, confirming that the dual moral framework requires the critical combination of higher education and direct experience of mobilization.

```{r}
#| label: fig-efectos-marginales
#| fig-cap: "Marginal effects of protest participation on violence justification by education"
#| fig-width: 14
#| fig-height: 10
#| echo: false
#| warning: false

# ===================================================================
# PANEL A: Marginal effect of participation by EDUCATION
# for PROTEST Violence
# ===================================================================

# Calculate marginal effects manually
# Marginal effect = β_protest + β_interaction * education_level

# Extract model coefficients
coef_educ_prot <- fixef(mod_educ_protesta_int)$cond

# Base marginal effect (secondary education = reference category)
me_base <- coef_educ_prot["protesta_dummy"]

# Marginal effects for each education level
me_data_educ_prot <- data.frame(
  education = c("High School\nor less", "Tech. College\nincomplete", "Tech. College\ncomplete",
                "University\nincomplete", "University\ncomplete"),
  me = c(
    me_base,  # Secondary (reference)
    me_base + coef_educ_prot["educ_cat_unorderedTechnical College incomplete:protesta_dummy"],
    me_base + coef_educ_prot["educ_cat_unorderedTechnical College complete:protesta_dummy"],
    me_base + coef_educ_prot["educ_cat_unorderedUniversity incomplete:protesta_dummy"],
    me_base + coef_educ_prot["educ_cat_unorderedUniversity complete:protesta_dummy"]
  )
)

# Calculate approximate standard errors using vcov
vcov_mat <- vcov(mod_educ_protesta_int)$cond

# SE for reference category (variance of protesta_dummy only)
se_base <- sqrt(vcov_mat["protesta_dummy", "protesta_dummy"])

# For interactions: SE = sqrt(Var(β1) + Var(β2) + 2*Cov(β1,β2))
se_tec_inc <- sqrt(
  vcov_mat["protesta_dummy", "protesta_dummy"] +
  vcov_mat["educ_cat_unorderedTechnical College incomplete:protesta_dummy",
           "educ_cat_unorderedTechnical College incomplete:protesta_dummy"] +
  2 * vcov_mat["protesta_dummy", "educ_cat_unorderedTechnical College incomplete:protesta_dummy"]
)

se_tec_comp <- sqrt(
  vcov_mat["protesta_dummy", "protesta_dummy"] +
  vcov_mat["educ_cat_unorderedTechnical College complete:protesta_dummy",
           "educ_cat_unorderedTechnical College complete:protesta_dummy"] +
  2 * vcov_mat["protesta_dummy", "educ_cat_unorderedTechnical College complete:protesta_dummy"]
)

se_univ_inc <- sqrt(
  vcov_mat["protesta_dummy", "protesta_dummy"] +
  vcov_mat["educ_cat_unorderedUniversity incomplete:protesta_dummy",
           "educ_cat_unorderedUniversity incomplete:protesta_dummy"] +
  2 * vcov_mat["protesta_dummy", "educ_cat_unorderedUniversity incomplete:protesta_dummy"]
)

se_univ_comp <- sqrt(
  vcov_mat["protesta_dummy", "protesta_dummy"] +
  vcov_mat["educ_cat_unorderedUniversity complete:protesta_dummy",
           "educ_cat_unorderedUniversity complete:protesta_dummy"] +
  2 * vcov_mat["protesta_dummy", "educ_cat_unorderedUniversity complete:protesta_dummy"]
)

me_data_educ_prot$se <- c(se_base, se_tec_inc, se_tec_comp, se_univ_inc, se_univ_comp)
me_data_educ_prot$ci_low <- me_data_educ_prot$me - 1.96 * me_data_educ_prot$se
me_data_educ_prot$ci_high <- me_data_educ_prot$me + 1.96 * me_data_educ_prot$se
me_data_educ_prot$significant <- !(me_data_educ_prot$ci_low < 0 & me_data_educ_prot$ci_high > 0)

# Reorder levels
me_data_educ_prot$education <- factor(me_data_educ_prot$education,
                                       levels = c("High School\nor less", "Tech. College\nincomplete",
                                                  "Tech. College\ncomplete", "University\nincomplete",
                                                  "University\ncomplete"))

# Panel A plot
panel_a <- ggplot(me_data_educ_prot, aes(x = education, y = me)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 0.8) +
  geom_point(aes(color = significant), size = 3.5) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high, color = significant),
                width = 0.2, linewidth = 1) +
  scale_color_manual(values = c("FALSE" = "gray50", "TRUE" = "#2C7BB6"),
                     guide = "none") +
  labs(title = "Protest Violence",
       x = "Education Level",
       y = "Marginal Effect of Participation\n(change in justification)") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 11),
        plot.title = element_text(size = 12, face = "bold"),
        panel.grid.minor = element_blank())

# ===================================================================
# PANEL B: Marginal effect of participation by EDUCATION
# for STATE Violence
# ===================================================================

coef_educ_est <- fixef(mod_educ_estatal_int)$cond
me_base_est <- coef_educ_est["protesta_dummy"]

me_data_educ_est <- data.frame(
  education = c("High School\nor less", "Tech. College\nincomplete", "Tech. College\ncomplete",
                "University\nincomplete", "University\ncomplete"),
  me = c(
    me_base_est,
    me_base_est + coef_educ_est["educ_cat_unorderedTechnical College incomplete:protesta_dummy"],
    me_base_est + coef_educ_est["educ_cat_unorderedTechnical College complete:protesta_dummy"],
    me_base_est + coef_educ_est["educ_cat_unorderedUniversity incomplete:protesta_dummy"],
    me_base_est + coef_educ_est["educ_cat_unorderedUniversity complete:protesta_dummy"]
  )
)

vcov_mat_est <- vcov(mod_educ_estatal_int)$cond
se_base_est <- sqrt(vcov_mat_est["protesta_dummy", "protesta_dummy"])

se_tec_inc_est <- sqrt(
  vcov_mat_est["protesta_dummy", "protesta_dummy"] +
  vcov_mat_est["educ_cat_unorderedTechnical College incomplete:protesta_dummy",
               "educ_cat_unorderedTechnical College incomplete:protesta_dummy"] +
  2 * vcov_mat_est["protesta_dummy", "educ_cat_unorderedTechnical College incomplete:protesta_dummy"]
)

se_tec_comp_est <- sqrt(
  vcov_mat_est["protesta_dummy", "protesta_dummy"] +
  vcov_mat_est["educ_cat_unorderedTechnical College complete:protesta_dummy",
               "educ_cat_unorderedTechnical College complete:protesta_dummy"] +
  2 * vcov_mat_est["protesta_dummy", "educ_cat_unorderedTechnical College complete:protesta_dummy"]
)

se_univ_inc_est <- sqrt(
  vcov_mat_est["protesta_dummy", "protesta_dummy"] +
  vcov_mat_est["educ_cat_unorderedUniversity incomplete:protesta_dummy",
               "educ_cat_unorderedUniversity incomplete:protesta_dummy"] +
  2 * vcov_mat_est["protesta_dummy", "educ_cat_unorderedUniversity incomplete:protesta_dummy"]
)

se_univ_comp_est <- sqrt(
  vcov_mat_est["protesta_dummy", "protesta_dummy"] +
  vcov_mat_est["educ_cat_unorderedUniversity complete:protesta_dummy",
               "educ_cat_unorderedUniversity complete:protesta_dummy"] +
  2 * vcov_mat_est["protesta_dummy", "educ_cat_unorderedUniversity complete:protesta_dummy"]
)

me_data_educ_est$se <- c(se_base_est, se_tec_inc_est, se_tec_comp_est, se_univ_inc_est, se_univ_comp_est)
me_data_educ_est$ci_low <- me_data_educ_est$me - 1.96 * me_data_educ_est$se
me_data_educ_est$ci_high <- me_data_educ_est$me + 1.96 * me_data_educ_est$se
me_data_educ_est$significant <- !(me_data_educ_est$ci_low < 0 & me_data_educ_est$ci_high > 0)

me_data_educ_est$education <- factor(me_data_educ_est$education,
                                      levels = c("High School\nor less", "Tech. College\nincomplete",
                                                 "Tech. College\ncomplete", "University\nincomplete",
                                                 "University\ncomplete"))

panel_b <- ggplot(me_data_educ_est, aes(x = education, y = me)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 0.8) +
  geom_point(aes(color = significant), size = 3.5) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high, color = significant),
                width = 0.2, linewidth = 1) +
  scale_color_manual(values = c("FALSE" = "gray50", "TRUE" = "#2C7BB6"),
                     guide = "none") +
  labs(title = "State Violence",
       x = "Education Level",
       y = "Marginal Effect of Participation\n(change in justification)") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 11),
        plot.title = element_text(size = 12, face = "bold"),
        panel.grid.minor = element_blank())

# Combine panels
combined_marginal_effects <- (panel_a | panel_b) +
  plot_annotation(
    tag_levels = 'A',
    caption = "Marginal effects with 95% CI. Red line indicates null effect. Grey points: non-significant.\nMarginal effect = change in justification from participating vs. not participating, conditioned by education. Compiled by the author based on ELSOC.",
    theme = theme(plot.caption = element_text(size = 9, hjust = 0))
  )

combined_marginal_effects
```

@fig-efectos-marginales visualizes the marginal effects of participation by education, allowing direct evaluation of the hypotheses about differential effects.

In Panel A, all education levels show significant positive effects on protest violence justification, with magnitudes ranging from +0.27 (secondary education) to +0.55 (technical college incomplete). Crucially, the legitimizing effect is stronger among technical college (+0.55) and university (+0.40) education than among secondary education (+0.27), showing that education amplifies (rather than reduces) the effect of participation, inverting the civilizing effect among those who mobilize.

Panel B shows that only university-educated individuals have significant negative effects on state violence justification (-0.20 to -0.22), while secondary and technical education show null effects (grey points on the red line). University-educated participants thus tend to legitimize protest violence (+0.40) while simultaneously rejecting state violence (-0.22).


### The Education "Paradox Effect"

```{r}
#| label: fig-paradoja-educacion
#| fig-cap: "Interaction analysis: Education × Protest Participation"
#| fig-width: 12
#| fig-height: 9
#| echo: false
#| warning: false

# Generate predictions from the education model with interactions
pred_df <- ggpredict(mod_educ_protesta_int,
                     terms = c("educ_cat_unordered", "protesta_dummy")) %>%
  as.data.frame() %>%
  rename(participation = group) %>%
  mutate(participation = factor(participation,
                                 levels = c("0", "1"),
                                 labels = c("Did not participate", "Participated")))

# Panel A: Predictions
panel_a <- ggplot(pred_df, aes(x = x, y = predicted,
                                color = participation, group = participation,
                                fill = participation)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              alpha = 0.2, color = NA) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("Did not participate" = "#F8766D", "Participated" = "#00BFC4")) +
  scale_fill_manual(values = c("Did not participate" = "#F8766D", "Participated" = "#00BFC4")) +
  labs(
    title = "Predicted Values",
    x = "Education Level",
    y = "Justification (predicted)",
    color = "Participation",
    fill = "Participation"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    legend.position = "bottom",
    legend.text = element_text(size = 11),
    plot.title = element_text(size = 13, face = "bold")
  )

# Panel B: Interaction magnitude
interaction_df <- pred_df %>%
  select(x, participation, predicted) %>%
  pivot_wider(names_from = participation, values_from = predicted) %>%
  mutate(Difference = `Participated` - `Did not participate`)

panel_b <- ggplot(interaction_df, aes(x = x, y = Difference)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 0.6) +
  geom_col(fill = "#4A4A4A", alpha = 0.7, width = 0.6) +
  geom_text(aes(label = round(Difference, 2)),
            vjust = ifelse(interaction_df$Difference > 0, -0.5, 1.5),
            size = 2.5) +
  labs(
    title = "Interaction Magnitude",
    x = "Education Level",
    y = "Difference\n(Participated - Did not participate)"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    plot.title = element_text(size = 13, face = "bold")
  )

# Combine
plot_final <- panel_a / panel_b +
  plot_annotation(
    tag_levels = 'A',
    caption = "Panel A shows effects by group; Panel B quantifies the interaction. Compiled by the author based on ELSOC 2016–2023.",
    theme = theme(plot.caption = element_text(size = 7, hjust = 0))
  )

# Display plot
plot_final
```

@fig-paradoja-educacion reveals the "educational paradox" predicted in hypotheses H1a and H1b. Panel A shows that education reduces violence justification among external observers (descending red line from ~1.95 at secondary education to ~1.75 at university complete), consistent with the civilizing effect. However, education inverts its effect among participants (ascending blue line from ~2.20 at secondary education to ~2.35 at technical college incomplete, then ~2.30 at university complete), showing that more educated individuals construct ideological frameworks that legitimize tactical violence in mobilization contexts. Panel B quantifies this inversion: the interaction magnitude is highest at technical college incomplete (+0.55) and university incomplete (+0.45), confirming that the "paradox" is most pronounced at mid-to-upper education levels where cultural capital and mobilization experience converge.

## Bidirectional Reconfiguration of Violence Justification (Protest Violence vs. State Violence)

The interaction models (Model 1b and Model 2b) reveal the bidirectional dynamic predicted in H2a–H2b (university-educated). H2a is fully confirmed: for protest violence, university education reduces justification among non-participants ($-0.13***$), but participation generates an increase ($+0.27***$) magnified by positive interactions with higher education (technical college incomplete: $+0.28*$; university complete: $+0.13**$), nullifying the civilizing effect. H2b is fully confirmed: for state violence, the pattern reverses completely: education alone ($+0.06$, ns) and participation alone ($-0.04$, ns) generate no effects, but their negative interactions produce significant rejection among university-educated individuals (incomplete: $-0.16*$; complete: $-0.18***$), showing that only the combination of university education + participation activates the critical framework toward repression.

The controls confirm the impact of the 2019 outbreak: moderate increase in protest justification ($+0.05*$) but collapse in state violence ($-0.40***$), radically transforming attitudes toward police repression. Ideology operates in opposite directions: right-wing ideology reduces justification of protests ($-0.03***$) but increases that of state violence ($+0.05***$). Women reject state violence more ($-0.13***$) with no differences in protests. This pattern confirms that the bidirectional reconfiguration predicted in H2a–H2b (legitimizing protests, rejecting repression) is complete only among those who combine university education with direct experience of mobilization, crystallizing the sophisticated dual moral framework.

```{r}
#| label: fig-bidireccional-tiempo
#| fig-cap: "Temporal evolution of bidirectional reconfiguration by education and participation (2016–2023)"
#| fig-width: 12
#| fig-height: 14
#| dpi: 300
#| echo: false
#| warning: false

# Get predictions for protest violence
pred_protestas_tiempo <- ggpredict(mod_educ_protesta_int,
                                   terms = c("year", "protesta_dummy", "educ_cat_unordered [High School or less, Technical College incomplete, Technical College complete, University complete]"),
                                   typical = "mean")

# Get predictions for state violence
pred_estatal_tiempo <- ggpredict(mod_educ_estatal_int,
                                 terms = c("year", "protesta_dummy", "educ_cat_unordered [High School or less, Technical College incomplete, Technical College complete, University complete]"),
                                 typical = "mean")

# Convert to dataframes
pred_protestas_df <- as.data.frame(pred_protestas_tiempo)
pred_protestas_df$violence_type <- "Protest Violence"
pred_protestas_df$participation <- factor(pred_protestas_df$group,
                                          labels = c("Did not participate", "Participated"))
pred_protestas_df$education <- pred_protestas_df$facet

pred_estatal_df <- as.data.frame(pred_estatal_tiempo)
pred_estatal_df$violence_type <- "State Violence"
pred_estatal_df$participation <- factor(pred_estatal_df$group,
                                        labels = c("Did not participate", "Participated"))
pred_estatal_df$education <- pred_estatal_df$facet

# Combine both datasets
pred_combined <- rbind(pred_protestas_df, pred_estatal_df)

# Create plot with facets by education and participation
ggplot(pred_combined, aes(x = x, y = predicted, color = violence_type, group = violence_type)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = violence_type),
              alpha = 0.15, color = NA) +
  facet_grid(participation ~ education,
             labeller = labeller(
               participation = c("Did not participate" = "Non-Participants",
                                "Participated" = "Participants")
             )) +
  scale_color_manual(values = c("Protest Violence" = "#EFC000FF",
                                 "State Violence" = "#0073C2FF")) +
  scale_fill_manual(values = c("Protest Violence" = "#EFC000FF",
                                "State Violence" = "#0073C2FF")) +
  geom_vline(xintercept = 2019, linetype = "dashed", color = "red", alpha = 0.5) +
  labs(
    title = "Bidirectional Reconfiguration: Temporal Evolution by Education and Participation",
    x = "Year",
    y = "Predicted Violence Justification",
    color = "Type of Violence",
    fill = "Type of Violence",
    caption = "Dashed red line: 2019 (social outbreak). Shaded areas: 95% CI. Other predictors at mean values.\nCompiled by the author based on ELSOC 2016–2023."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 10),
    strip.text = element_text(face = "bold", size = 11),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    plot.caption = element_text(size = 8)
  ) +
  ylim(1.3, 2.4)
```

@fig-bidireccional-tiempo visualizes the temporal dynamics of the bidirectional reconfiguration predicted in H2a–H2b, showing that the dual moral framework requires both education and participation. Among non-participants (top row), the yellow (protests) and blue (state) lines remain parallel and close across all education levels, confirming that without participation there is no bidirectional reconfiguration despite the university civilizing effect (H1a operates but H2b is not activated).

Among participants (bottom row), three patterns emerge for evaluating H2a–H2b. Secondary/technical education shows partial rejection of H2a–H2b: elevated protest violence justification (+0.27 to +0.55, confirming H2a) but without significant rejection of state violence (rejecting H2b), evidencing tactical legitimization without a critical framework toward repression. Technical college incomplete reaches the highest radicalization on protests (>2.2 in 2019, $+0.28*$) but maintains parallel lines, confirming that technical capital without university socialization does not generate the normative differentiation predicted in H2b.

The complete dual framework predicted in H2a–H2b is confirmed exclusively among university-educated participants: (1) a sustained elevated yellow line (~2.0–2.1), nullifying the civilizing effect and confirming H2a; (2) a dramatic collapse of the blue line in 2019 from ~1.75 to ~1.35, the lowest level in the figure ($-0.40***$), confirming H2b; (3) maximum post-2019 divergence with a gap of ~0.7 points, evidencing the permanence of the dual framework. This pattern confirms that H2a and H2b are jointly satisfied only under the critical combination of university education + mobilization experience, crystallizing the sophisticated dual normative framework.
