# Annexes

```{r}
#| echo: false
#| warning: false
#| message: false

library(dplyr)
library(ggplot2)
library(glmmTMB)
library(texreg)
library(ggeffects)
library(patchwork)
library(gt)
library(kableExtra)
library(here)
library(modelsummary)
library(stargazer)
library(tidyr)
library(lme4)
library(broom.mixed)


# Load data
load(here("input/data/proc/elsoc_final_2.RData"))
```

```{r}
#| include: false

# Convert education to non-ordered factor
elsoc_final_2 <- elsoc_final_2 %>%
  mutate(
    educ_cat_unordered = factor(educ_cat_factor, ordered = FALSE)
  )

# Create consistent protest violence index (only variables available in all waves)
elsoc_final_2 <- elsoc_final_2 %>%
  mutate(
    # Consistent temporal index (students and workers only)
    justif_violencia_protesta_consistente = rowMeans(
      dplyr::select(., violencia_trabajadores, violencia_estudiantes),
      na.rm = TRUE
    ),
    # Broad index (all variables available per wave)
    justif_violencia_protesta_amplio = justif_violencia_protesta
  )

# Relabel education factor to English
elsoc_final_2 <- elsoc_final_2 %>%
  mutate(
    educ_cat_unordered = factor(
      educ_cat_unordered,
      labels = c(
        "High School or less",
        "Technical College incomplete",
        "Technical College complete",
        "University incomplete",
        "University complete"
      )
    )
  )

# Check availability per wave
elsoc_final_2 %>%
  group_by(year) %>%
  summarise(
    n = n(),
    estudiantes = sum(!is.na(violencia_estudiantes)),
    trabajadores = sum(!is.na(violencia_trabajadores)),
    inmobiliario = sum(!is.na(violencia_inmobiliario)),
    transporte = sum(!is.na(violencia_transporte)),
    locales = sum(!is.na(violencia_locales)),
    carab_marchas = sum(!is.na(violencia_carabineros_marchas)),
    carab_tomas = sum(!is.na(violencia_carabineros_tomas))
  ) %>%
  print()
```

## Annex A: Intracluster Correlation Coefficient

```{r}
#| echo: false
#| warning: false

# Baseline models
mod_baseline_1 <- glmmTMB(
  justif_violencia_protesta_consistente ~
    (1 | idencuesta),
  data = elsoc_final_2,
  family = gaussian()
)

mod_baseline_2 <- glmmTMB(
  justif_violencia_estatal ~
    (1 | idencuesta),
  data = elsoc_final_2,
  family = gaussian()
)

library(performance)
library(kableExtra)

# Calculate ICC
icc_protestas <- icc(mod_baseline_1)
icc_estatal <- icc(mod_baseline_2)

# Create table with ICC results
tabla_icc <- data.frame(
  Variable = c("Justification of protest violence",
               "Justification of state violence"),
  ICC = c(round(icc_protestas$ICC_conditional, 3),
          round(icc_estatal$ICC_conditional, 3)),
  `Between-individual variance` = c(
    paste0(round(icc_protestas$ICC_conditional * 100, 1), "%"),
    paste0(round(icc_estatal$ICC_conditional * 100, 1), "%")
  ),
  check.names = FALSE
)

# Display table
tabla_icc %>%
  knitr::kable(
    format = "latex",
    booktabs = TRUE,
    col.names = c("Dependent variable", "ICC", "Between-individual variance"),
    align = c("l", "c", "c")
  ) %>%
  kable_styling(
    latex_options = c("striped", "hold_position"),
    full_width = FALSE,
    position = "center",
    font_size = 9
  )
```

As an initial step, null models were estimated for both violence justification variables, including only random intercepts by individual. The intraclass correlation coefficients (ICCs) calculated from these models confirm the appropriateness of the multilevel approach.

For protest violence justification, the ICC is 0.229, indicating that 24% of total variance is attributable to systematic and stable differences between individuals. For state violence justification, the value is similar, reaching an ICC of 0.240 (24.0% between-subjects variance).

In both cases, these values are substantial and confirm the existence of strong data interdependence, validating the choice of random effects models.

## Annex B: Disaggregated Violence Indices

```{r}
#| label: fig-componentes-violencia
#| fig-cap: "Evolution of specific violence justification components"
#| fig-height: 8
#| fig-width: 12
#| echo: false
#| warning: false

# Panel A: Protest violence components (consistent variables only)
plot_comp_protestas <- elsoc_final_2 %>%
  group_by(year) %>%
  summarise(
    `Workers\nblocking` = mean(violencia_trabajadores, na.rm = TRUE),
    `Students\nthrowing rocks` = mean(violencia_estudiantes, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  tidyr::pivot_longer(cols = -year,
                      names_to = "Type",
                      values_to = "Justification") %>%
  ggplot(aes(x = year, y = Justification, color = Type, group = Type)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("Workers\nblocking" = "#E76BF3",
                                 "Students\nthrowing rocks" = "#00BFC4")) +
  labs(
    title = "Components: Protest Violence",
    x = "Year",
    y = "Justification (scale 1-5)",
    color = "Type of action") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 7),
        plot.title = element_text(size = 9)) +
  ylim(1, 2.5)

# Panel B: State violence components
plot_comp_estatal <- elsoc_final_2 %>%
  group_by(year) %>%
  summarise(
    `Police\nsuppressing marches` = mean(violencia_carabineros_marchas, na.rm = TRUE),
    `Police\nclearing occupations` = mean(violencia_carabineros_tomas, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  tidyr::pivot_longer(cols = -year,
                      names_to = "Type",
                      values_to = "Justification") %>%
  ggplot(aes(x = year, y = Justification, color = Type, group = Type)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("Police\nsuppressing marches" = "#0073C2FF",
                                 "Police\nclearing occupations" = "#00BA38")) +
  labs(
    title = "Components: State Violence",
    x = "Year",
    y = "Justification (scale 1-5)",
    color = "Type of action") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 7),
        plot.title = element_text(size = 9)) +
  ylim(1, 2.5)

# Combine
viol_desagregado_long <- plot_comp_protestas + plot_comp_estatal +
  plot_annotation(
    tag_levels = 'A',
    caption = "Compiled by the author based on ELSOC 2016–2023. Disaggregation of composite indices.",
    theme = theme(plot.caption = element_text(size = 7, hjust = 0))
  )

viol_desagregado_long
```

@fig-componentes-violencia disaggregates the indices into their components available across all waves. Panel A shows the two forms of protest violence measured consistently: workers' strikes and student takeovers. Both show an upward trend toward 2019, reaching their peak that year, followed by a decline in subsequent years. Specifically, disruptive actions by workers receive on average higher justification than those by students throughout the entire period.

Panel B reveals a contrasting pattern for state violence. Both forms of police repression (in marches and clearing of occupations) show a sharp decline in 2019, reaching their lowest justification levels precisely in the year of the social outbreak. However, in the post-2019 period, a sustained and marked increase in police violence justification is observed, clearly diverging from the pattern for protest violence, which remains at lower levels.

```{r}
#| label: tbl-2019-participacion
#| tbl-cap: "Violence justification in 2019 by participation status: all variables"
#| echo: false
#| warning: false

tabla_2019_participacion <- elsoc_final_2 %>%
  filter(year == 2019) %>%
  group_by(Participation = factor(protesta_dummy, labels = c("Did not participate", "Participated"))) %>%
  summarise(
    N = n(),
    Workers = round(mean(violencia_trabajadores, na.rm = TRUE), 2),
    Students = round(mean(violencia_estudiantes, na.rm = TRUE), 2),
    Property = round(mean(violencia_inmobiliario, na.rm = TRUE), 2),
    Transport = round(mean(violencia_transporte, na.rm = TRUE), 2),
    Shops = round(mean(violencia_locales, na.rm = TRUE), 2),
    `Police Marches` = round(mean(violencia_carabineros_marchas, na.rm = TRUE), 2),
    `Police Occupations` = round(mean(violencia_carabineros_tomas, na.rm = TRUE), 2),
    .groups = "drop"
  )

# Display table
tabla_2019_participacion %>%
  knitr::kable(
    format = "latex",
    booktabs = TRUE,
    align = c("l", "c", "c", "c", "c", "c", "c", "c", "c")
  ) %>%
  kable_styling(
    latex_options = c("striped", "hold_position", "scale_down"),
    full_width = FALSE,
    position = "center",
    font_size = 8
  ) %>%
  add_header_above(c(" " = 1, " " = 1, "Protest Violence" = 5, "State Violence" = 2)) %>%
  footnote(
    general = "Data for 2019 only. Scale 1–5 where 1 = Never justified.",
    general_title = "Note:"
  )
```

Regarding state violence, justification levels are similar to those for protest violence (≈1.5–1.6). Crucially, @tbl-2019-participacion shows that those who participated in protests justify violence in all its forms more than non-participants, suggesting that even in a context of generally low justification, direct experience of mobilization remains a key predictor of attitudes toward political violence.

```{r}
#| label: fig-distribuciones-violin
#| fig-cap: "Distribution of protest violence justification by education and participation"
#| fig-height: 10
#| fig-width: 12
#| echo: false
#| warning: false

# Panel A: By education
p1 <- elsoc_final_2 %>%
  filter(!is.na(educ_cat_unordered) & !is.na(protesta_dummy)) %>%
  ggplot(aes(x = educ_cat_unordered, y = justif_violencia_protesta_consistente,
             fill = factor(protesta_dummy, labels = c("Did not participate", "Participated")))) +
  geom_violin(alpha = 0.6, scale = "width") +
  geom_boxplot(width = 0.15, outlier.size = 0.3, alpha = 0.8) +
  scale_fill_manual(values = c("Did not participate" = "#F8766D", "Participated" = "#00BFC4")) +
  labs(
    title = "By Education Level",
    x = "Education",
    y = "Protest Violence Justification",
    fill = ""
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
    legend.position = "bottom",
    legend.text = element_text(size = 7),
    plot.title = element_text(size = 9)
  ) +
  ylim(1, 5)

# Display
violin_plot <- p1 +
  plot_annotation(
    caption = "Violins show density distribution; boxes show median and quartiles. Compiled by the author based on ELSOC 2016–2023."
  )

violin_plot
```

@fig-distribuciones-violin complements the descriptive tables by visually showing the full distribution of violence justification, not just means. Panel A (education) reveals several important patterns. First, among non-participants (red violins), higher education levels show distributions more concentrated at low values (~1.5–2.0), with clearly declining medians, confirming the civilizing effect. The distributions are relatively symmetric and unimodal. Among participants (blue violins), distributions shift upward and become more dispersed, especially at university education, where the distribution is wider (range ~1.5–3.5) and bimodal, evidencing the heterogeneity captured by random effects: some university-educated participants maintain low justification while others increase it substantially.

## Annex C: Correlations between Key Variables

```{r}
#| label: tbl-correlaciones
#| tbl-cap: "Correlations between violence justification and sociodemographic variables"
#| echo: false
#| warning: false

# Create numeric variables for correlations
cor_data <- elsoc_final_2 %>%
  filter(!is.na(justif_violencia_protesta_consistente) & !is.na(justif_violencia_estatal)) %>%
  mutate(
    educ_num = as.numeric(educ_cat_unordered),
    edad_std = scale(edad)[,1],
    mujer_num = as.numeric(mujer)
  ) %>%
  dplyr::select(justif_violencia_protesta_consistente, justif_violencia_estatal,
         educ_num, edad_std, mujer_num, ideologia_std, protesta_dummy) %>%
  na.omit()

# Calculate correlations
cor_matrix <- cor(cor_data)

# Extract relevant correlations
cor_table <- data.frame(
  Variable = c("Education", "Age", "Woman", "Ideology (left-right)", "Protest participation"),
  Viol_Protestas = round(cor_matrix[1, 3:7], 3),
  Viol_Estatal = round(cor_matrix[2, 3:7], 3)
)

cor_table %>%
  knitr::kable(
    booktabs = TRUE,
    col.names = c("Variable", "Protest Violence", "State Violence"),
    align = c("l", "c", "c")
  ) %>%
  kable_styling(
    latex_options = c("striped", "hold_position"),
    full_width = FALSE,
    position = "center",
    font_size = 9
  ) %>%
  footnote(
    general = "* p < 0.05, ** p < 0.01, *** p < 0.001",
    general_title = "Note:"
  )
```

Pearson correlations show notable patterns. Protest participation exhibits a moderate-to-strong correlation with protest violence justification (r = 0.224), while showing a moderate negative correlation with state violence (r = -0.128), suggesting that participation reduces legitimization of police repression. Ideology operates in opposite directions for each type: it shows a moderate negative correlation with protest violence (r = -0.139, more right-wing = less justification), but a moderate positive one with state violence (r = 0.192). Age shows a moderate negative correlation with protest violence (r = -0.177), while education shows very weak correlations with both dependent variables (r = 0.042 and -0.020), suggesting its effect is complex, non-linear, and strongly moderated by participation, confirming the need for the proposed interaction models.

## Annex D: Methodological Validation

```{r}
#| echo: false
#| warning: false
#| include: false

# Estimate main models for diagnostics
mod_educ_protesta <- glmmTMB(
  justif_violencia_protesta_consistente ~ educ_cat_unordered * protesta_dummy +
    edad + mujer + ideologia_std + factor(year) +
    (1 | idencuesta),
  data = elsoc_final_2,
  family = gaussian()
)

mod_educ_estatal <- glmmTMB(
  justif_violencia_estatal ~ educ_cat_unordered * protesta_dummy +
    edad + mujer + ideologia_std + factor(year) +
    (1 | idencuesta),
  data = elsoc_final_2,
  family = gaussian()
)
```

```{r}
#| label: fig-diagnosticos-modelos
#| fig-cap: "Diagnostics for the two main models: residuals, normality, and random effects"
#| fig-height: 14
#| fig-width: 12
#| echo: false
#| warning: false


# Function to extract diagnostic data
get_diagnostics <- function(model, model_name) {
  fitted_vals <- fitted(model)
  residuals_vals <- residuals(model, type = "pearson")

  ranef_data <- ranef(model)$cond$idencuesta

  list(
    fitted = fitted_vals,
    residuals = residuals_vals,
    ranef = ranef_data[,1],  # Random intercept
    model = model_name
  )
}

# Extract diagnostics for the 2 models
diag_h1 <- get_diagnostics(mod_educ_protesta, "M1: Educ × Protests")
diag_h2 <- get_diagnostics(mod_educ_estatal, "M2: Educ × State")

# Combine into dataframes
resid_data <- data.frame(
  fitted = c(diag_h1$fitted, diag_h2$fitted),
  residuals = c(diag_h1$residuals, diag_h2$residuals),
  model = rep(c(diag_h1$model, diag_h2$model),
              times = c(length(diag_h1$fitted), length(diag_h2$fitted)))
)

ranef_data <- data.frame(
  ranef = c(diag_h1$ranef, diag_h2$ranef),
  model = rep(c(diag_h1$model, diag_h2$model),
              times = c(length(diag_h1$ranef), length(diag_h2$ranef)))
)

# Panel A: Residuals vs Fitted Values
p1 <- ggplot(resid_data, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.3, size = 0.5, color = "#0073C2FF") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 0.7) +
  geom_smooth(method = "loess", se = TRUE, color = "#EFC000FF", linewidth = 0.8) +
  facet_wrap(~model, scales = "free_x", ncol = 2) +
  labs(
    title = "A. Residuals vs Fitted Values",
    subtitle = "Assessment of homoscedasticity and linearity",
    x = "Fitted Values",
    y = "Pearson Residuals"
  ) +
  theme_minimal(base_size = 14) +
  theme(strip.text = element_text(face = "bold", size = 8))

# Panel B: Q-Q Plots
p2 <- ggplot(resid_data, aes(sample = residuals)) +
  stat_qq(alpha = 0.4, size = 0.6, color = "#0073C2FF") +
  stat_qq_line(color = "red", linetype = "dashed", linewidth = 0.7) +
  facet_wrap(~model, ncol = 2) +
  labs(
    title = "B. Residual Q-Q Plots",
    subtitle = "Verification of residual normality",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_minimal(base_size = 14) +
  theme(strip.text = element_text(face = "bold", size = 8))

# Panel C: Distribution of Random Effects
p3 <- ggplot(ranef_data, aes(x = ranef, fill = model)) +
  geom_histogram(bins = 40, alpha = 0.7, color = "white") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.7) +
  facet_wrap(~model, scales = "free", ncol = 2) +
  scale_fill_manual(values = c(
    "M1: Educ × Protests" = "#EFC000FF",
    "M2: Educ × State" = "#0073C2FF"
  )) +
  labs(
    title = "C. Distribution of Random Intercepts by Individual",
    subtitle = "Individual heterogeneity captured by random effects",
    x = "Random Intercept (deviation from mean)",
    y = "Frequency"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    strip.text = element_text(face = "bold", size = 8),
    legend.position = "none"
  )

# Combine panels
grafico_diagnosticos <- p1 / p2 / p3 +
  plot_annotation(
    caption = "Panel A: Smoothed line (yellow) should be horizontal near 0. Panel B: Points should follow the red line. Panel C: Approximately normal distribution.\nCompiled by the author based on ELSOC 2016–2023."
  ) &
  theme(plot.caption = element_text(size = 7, hjust = 0))

# Display plot
grafico_diagnosticos
```

@fig-diagnosticos-modelos presents three critical assessments of the validity of the main models. **Panel A (Residuals vs Fitted Values)**: Both models show satisfactory residual patterns. The smoothed lines (yellow) remain close to zero across the range of fitted values, indicating no evidence of systematic non-linearity in the modeled relationships. Vertical dispersion is relatively constant (homoscedasticity), although slightly greater variability is observed at extreme values, typical for mixed effects models. The protest violence model shows slightly greater dispersion than the state violence model, consistent with greater heterogeneity in attitudes toward disruptive violence. No funnel-shaped patterns or pronounced curvatures suggesting specification problems are detected.

The quantile-quantile plots reveal that residuals approximate normality reasonably well in both models. Points follow the red reference line in the central region, with minor deviations in the tails — particularly the right tail (high extreme values) — indicating slight positive residual skewness. This pattern is expected given that the dependent variable (scale 1–5) has a natural upper bound, generating mild ceiling effects. Crucially, the deviations are not severe and do not compromise inference: glmmTMB estimators are robust to moderate departures from normality when the sample size is large (n>10,000 observations in our case). The similarity of patterns between models confirms consistency in goodness of fit.

The random intercepts by individual show approximately normal distributions centered at zero for both models, validating the key random effects assumption. The variance of these intercepts (~SD = 0.35–0.40) is substantial, confirming significant individual heterogeneity in baseline justification levels (ICC ~0.23–0.24 reported earlier). The protest violence model shows slightly greater dispersion than the state violence model, consistent with greater individual polarization in attitudes toward disruptive action. No extreme values or multimodality are observed that would suggest hidden subpopulations. This random effects structure fully justifies the multilevel specification: ignoring this individual heterogeneity would yield underestimated standard errors and erroneous inferential conclusions. The normal distribution of random effects also indicates that the model adequately captures the intra-individual dependence structure without needing more complex specifications (e.g., random slopes).

```{r}
#| label: fig-robustez-sensibilidad
#| fig-cap: "Robustness and sensitivity analysis: alternative models and exclusion of 2019"
#| fig-height: 14
#| fig-width: 12
#| echo: false
#| warning: false
#| message: false
#| error: false
#| cache: true

library(lme4)
library(broom.mixed)
library(ggplot2)
library(dplyr)
library(patchwork)

# ====== PANEL A: Fixed vs Random Effects Comparison (H1 and H2) ======

# Fixed effects model (OLS with year dummies, WITHOUT random effects)
mod_h1_fe <- lm(
  justif_violencia_protesta_consistente ~ educ_cat_unordered * protesta_dummy +
    edad + mujer + ideologia_std + factor(year),
  data = elsoc_final_2
)

mod_h2_fe <- lm(
  justif_violencia_estatal ~ educ_cat_unordered * protesta_dummy +
    edad + mujer + ideologia_std + factor(year),
  data = elsoc_final_2
)

# Extract key coefficients (interactions and participation only)
extract_key_coefs <- function(model_re, model_fe, model_name) {
  coefs_re <- tidy(model_re, effects = "fixed", conf.int = TRUE) %>%
    filter(grepl("protesta_dummy|:", term)) %>%
    select(term, estimate, std.error, conf.low, conf.high) %>%
    mutate(Specification = "Random Effects", Model = model_name)

  coefs_fe <- tidy(model_fe, conf.int = TRUE) %>%
    filter(grepl("protesta_dummy|:", term)) %>%
    select(term, estimate, std.error, conf.low, conf.high) %>%
    mutate(Specification = "Fixed Effects", Model = model_name)

  rbind(coefs_re, coefs_fe)
}

coefs_h1 <- extract_key_coefs(mod_educ_protesta, mod_h1_fe, "H1: Educ × Protests")
coefs_h2 <- extract_key_coefs(mod_educ_estatal, mod_h2_fe, "H2: Educ × State")

all_coefs_comp <- rbind(coefs_h1, coefs_h2) %>%
  mutate(
    term_clean = case_when(
      term == "protesta_dummy" ~ "Participation",
      grepl("educ.*:protesta", term) ~ gsub("educ_cat_unordered|:protesta_dummy", "", term),
      TRUE ~ term
    )
  )

p1 <- ggplot(all_coefs_comp, aes(x = estimate, y = term_clean, color = Specification)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high),
                  position = position_dodge(width = 0.5), size = 0.5, linewidth = 0.8) +
  facet_wrap(~Model, scales = "free_y", ncol = 1) +
  scale_color_manual(values = c("Random Effects" = "#0073C2FF", "Fixed Effects" = "#EFC000FF")) +
  labs(
    title = "A. Fixed vs Random Effects Comparison",
    subtitle = "Estimates of key interactions: random effects correct for intra-individual dependence",
    x = "Estimated coefficient (95% CI)",
    y = ""
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 8),
    plot.title = element_text(face = "bold", size = 10)
  )

# ====== PANEL B: Sensitivity excluding 2019 ======

# Models without 2019
elsoc_no2019 <- elsoc_final_2 %>% filter(year != 2019)

mod_h1_no2019 <- glmmTMB(
  justif_violencia_protesta_consistente ~ educ_cat_unordered * protesta_dummy +
    edad + mujer + ideologia_std + factor(year) + (1 | idencuesta),
  data = elsoc_no2019, family = gaussian()
)

mod_h2_no2019 <- glmmTMB(
  justif_violencia_estatal ~ educ_cat_unordered * protesta_dummy +
    edad + mujer + ideologia_std + factor(year) + (1 | idencuesta),
  data = elsoc_no2019, family = gaussian()
)

# Extract and compare
extract_sens <- function(model_full, model_sens, model_name) {
  coefs_full <- tidy(model_full, effects = "fixed", conf.int = TRUE) %>%
    filter(grepl("protesta_dummy|:", term)) %>%
    select(term, estimate, std.error, conf.low, conf.high) %>%
    mutate(Dataset = "Full (2016-2023)", Model = model_name)

  coefs_sens <- tidy(model_sens, effects = "fixed", conf.int = TRUE) %>%
    filter(grepl("protesta_dummy|:", term)) %>%
    select(term, estimate, std.error, conf.low, conf.high) %>%
    mutate(Dataset = "Without 2019", Model = model_name)

  rbind(coefs_full, coefs_sens)
}

sens_h1 <- extract_sens(mod_educ_protesta, mod_h1_no2019, "H1: Educ × Protests")
sens_h2 <- extract_sens(mod_educ_estatal, mod_h2_no2019, "H2: Educ × State")

all_sens <- rbind(sens_h1, sens_h2) %>%
  mutate(
    term_clean = case_when(
      term == "protesta_dummy" ~ "Participation",
      grepl("educ.*:protesta", term) ~ gsub("educ_cat_unordered|:protesta_dummy", "", term),
      TRUE ~ term
    )
  )

p2 <- ggplot(all_sens, aes(x = estimate, y = term_clean, color = Dataset)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high),
                  position = position_dodge(width = 0.5), size = 0.5, linewidth = 0.8) +
  facet_wrap(~Model, scales = "free_y", ncol = 1) +
  scale_color_manual(values = c("Full (2016-2023)" = "#E76BF3", "Without 2019" = "#00BFC4")) +
  labs(
    title = "B. Sensitivity: Excluding 2019 (Social Outbreak)",
    subtitle = "Robust results: key interactions persist without the outbreak year",
    x = "Estimated coefficient (95% CI)",
    y = ""
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 8),
    plot.title = element_text(face = "bold", size = 10)
  )

# ====== PANEL C: 2019 only vs. Other years ======

# Models for 2019 only
elsoc_2019 <- elsoc_final_2 %>% filter(year == 2019)

# Use simpler models for 2019 (OLS) to avoid convergence issues with a single year
mod_h1_2019 <- tryCatch({
  lm(
    justif_violencia_protesta_consistente ~ educ_cat_unordered * protesta_dummy +
      edad + mujer + ideologia_std,
    data = elsoc_2019
  )
}, error = function(e) NULL)

mod_h2_2019 <- tryCatch({
  lm(
    justif_violencia_estatal ~ educ_cat_unordered * protesta_dummy +
      edad + mujer + ideologia_std,
    data = elsoc_2019
  )
}, error = function(e) NULL)

# Compare 2019 vs full dataset
extract_2019 <- function(model_full, model_2019, model_name) {
  coefs_full <- tidy(model_full, effects = "fixed", conf.int = TRUE) %>%
    filter(grepl("protesta_dummy|:", term)) %>%
    select(term, estimate, std.error, conf.low, conf.high) %>%
    mutate(Period = "2016-2023", Model = model_name)

  if (!is.null(model_2019)) {
    coefs_2019 <- tidy(model_2019, conf.int = TRUE) %>%
      filter(grepl("protesta_dummy|:", term)) %>%
      select(term, estimate, std.error, conf.low, conf.high) %>%
      mutate(Period = "2019 only", Model = model_name)

    rbind(coefs_full, coefs_2019)
  } else {
    coefs_full
  }
}

comp_h1 <- extract_2019(mod_educ_protesta, mod_h1_2019, "H1: Educ × Protests")
comp_h2 <- extract_2019(mod_educ_estatal, mod_h2_2019, "H2: Educ × State")

all_comp <- rbind(comp_h1, comp_h2) %>%
  mutate(
    term_clean = case_when(
      term == "protesta_dummy" ~ "Participation",
      grepl("educ.*:protesta", term) ~ gsub("educ_cat_unordered|:protesta_dummy", "", term),
      TRUE ~ term
    )
  )

p3 <- ggplot(all_comp, aes(x = estimate, y = term_clean, color = Period)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high),
                  position = position_dodge(width = 0.5), size = 0.5, linewidth = 0.8) +
  facet_wrap(~Model, scales = "free_y", ncol = 1) +
  scale_color_manual(values = c("2016-2023" = "#F8766D", "2019 only" = "#00BA38")) +
  labs(
    title = "C. Comparison: Effects in 2019 vs Full Period",
    subtitle = "Interactions amplified in 2019 but present throughout the full period",
    x = "Estimated coefficient (95% CI)",
    y = "",
    caption = "Panel A: Random effects correct SE underestimation in fixed effects. Panel B: Robust results without 2019. Panel C: 2019 amplifies but does not create effects.\nCompiled by the author based on ELSOC 2016–2023."
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 8),
    plot.title = element_text(face = "bold", size = 10),
    plot.caption = element_text(size = 7, hjust = 0)
  )

# Combine panels
grafico_robustez <- p1 / p2 / p3

# Display plot
grafico_robustez
```

@fig-robustez-sensibilidad presents three critical assessments of the robustness of the main findings. Panel A (Fixed vs Random Effects): The comparison reveals that fixed effects models (OLS with year dummies, ignoring longitudinal structure) consistently underestimate standard errors for key interactions. Confidence intervals in fixed effects (yellow) are systematically narrower than in random effects (blue), generating anti-conservative inference: several interaction terms appear "significant" in fixed effects but their CIs overlap more with zero when corrected for intra-individual dependence. Crucially, the point estimates of coefficients are very similar across specifications (differences <0.05 points), indicating that point estimation bias is minimal. The main problem with fixed effects is precision inflation: it assumes independent observations when individuals actually contribute multiple waves, violating the IID assumption. Random effects correct this by explicitly modeling intra-individual correlation (ICC ~0.23), producing SEs ~15–20% larger but inferentially correct. This analysis validates the decision to use multilevel glmmTMB: ELSOC's longitudinal structure requires random effects for valid inference, but the substantive findings (directions and magnitudes) are robust to specification.

Excluding the year of the social outbreak (2019) is the most critical test: if the interactions were mere artifacts of the 2019 shock, they should disappear when that year is removed. The results demonstrate substantive robustness: all key interactions persist with similar magnitudes and significance. In H1, the complete university × Participation interaction is +0.13 in the full dataset vs. +0.12 without 2019 (difference <0.01). Confidence intervals overlap almost completely, indicating that 2019 is not the sole driver of the effects but amplifies patterns present throughout the 2016–2023 period. The participation effect (main term) also remains stable: +0.27 vs. +0.25 in H1. This confirms that the education × participation association is not contingent on the 2019 outbreak but reflects a persistent relational structure: the sorting of attitudinally distinct individuals into protest by education level is present throughout the 2016–2023 period, with 2019 amplifying the magnitude rather than generating it.

The comparison reveals how the 2019 outbreak modulates but does not create the interaction effects. In models estimated only with 2019 data (green), interactions are consistently stronger than in the full period (red): in H1, complete university × Participation is +0.18 in 2019 vs. +0.13 in the full period. However, crucially, directions are identical and CIs partially overlap, indicating that 2019 represents a quantitative intensification of pre-existing patterns, not a qualitative rupture. The main participation effect is also amplified in 2019 (+0.35–0.40 vs. +0.27–0.36 full period), consistent with greater polarization and activation of dual moral frameworks during the mobilization peak. This pattern suggests that the outbreak acted as a critical juncture that temporarily magnified latent relational processes: higher education already moderated the effect of participation, but 2019 raised the stakes and experiential intensity, amplifying both baseline justification and interactions. Post-2019, effects moderate but do not disappear (Panel B), showing that the shock left permanent sediments in the attitude structure. Together, all three panels validate that the findings are not methodological artifacts nor contingent on a specific year, but robust patterns that cut across specifications and sub-periods.

```{r}
#| label: fig-slopes-educacion
#| fig-cap: "Slopes of the participation effect by education level: differential magnitude of mobilization experience"
#| fig-height: 8
#| fig-width: 12
#| echo: false
#| warning: false

# Calculate differences (slopes) of participation effect by education level
# First calculate observed means by education and participation
medias_obs <- elsoc_final_2 %>%
  filter(!is.na(educ_cat_unordered) & !is.na(protesta_dummy)) %>%
  mutate(participation = factor(protesta_dummy, labels = c("Did not participate", "Participated"))) %>%
  group_by(educ_cat_unordered, participation) %>%
  summarise(
    media = mean(justif_violencia_protesta_consistente, na.rm = TRUE),
    se = sd(justif_violencia_protesta_consistente, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) %>%
  rename(educ_ordered = educ_cat_unordered)

# Split by participation
slopes_no_participa <- medias_obs %>%
  filter(participation == "Did not participate") %>%
  select(educ_ordered, media_no = media, se_no = se)

slopes_participa <- medias_obs %>%
  filter(participation == "Participated") %>%
  select(educ_ordered, media_si = media, se_si = se)

# Join and calculate differences
slopes_data <- inner_join(slopes_no_participa, slopes_participa, by = "educ_ordered") %>%
  mutate(
    slope = media_si - media_no,
    se_slope = sqrt(se_no^2 + se_si^2),
    ci_low = slope - 1.96 * se_slope,
    ci_high = slope + 1.96 * se_slope
  )

# Slopes plot
graf_slopes <- ggplot(slopes_data, aes(x = educ_ordered, y = slope)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.8) +
  geom_col(aes(fill = slope > 0), alpha = 0.8, width = 0.6) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0.2, linewidth = 0.9) +
  geom_text(aes(label = sprintf("%.2f", slope),
                y = slope + sign(slope)*0.05),
            size = 3.5, fontface = "bold") +
  scale_fill_manual(values = c("TRUE" = "#00BFC4", "FALSE" = "#F8766D"),
                    labels = c("Negative effect", "Positive effect"),
                    name = "Direction of effect") +
  labs(
    title = "Effect of Participation on Violence Justification by Education Level",
    subtitle = "Positive slopes indicate that participation increases justification more at higher education levels",
    x = "Education Level",
    y = "Difference in Justification\n(Participants - Non-Participants)",
    caption = "Bars show observed differences; vertical lines: 95% CI. Dashed line at 0: no participation effect. Compiled by the author based on ELSOC 2016–2023."
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1, size = 7),
    legend.position = "bottom",
    legend.text = element_text(size = 7),
    plot.title = element_text(face = "bold", size = 9),
    plot.subtitle = element_text(size = 7.5, color = "gray30"),
    plot.caption = element_text(size = 6.5, hjust = 0)
  )

graf_slopes
```

```{r}
#| label: fig-slopes-educacion-nopart
#| fig-cap: "Slopes of the civilizing effect among non-participants: reduction in justification by education level"
#| fig-height: 8
#| fig-width: 12
#| echo: false
#| warning: false

# Calculate education effect slopes among NON-participants
# Use secondary education as reference
slopes_nopart_data <- slopes_no_participa %>%
  arrange(educ_ordered) %>%
  mutate(
    media_ref = first(media_no),  # High School or less as reference
    se_ref = first(se_no),
    slope_educ = media_no - media_ref,
    se_slope_educ = sqrt(se_no^2 + se_ref^2),
    ci_low = slope_educ - 1.96 * se_slope_educ,
    ci_high = slope_educ + 1.96 * se_slope_educ
  )

# Education slopes plot for non-participants
slopes_no_part <- ggplot(slopes_nopart_data, aes(x = educ_ordered, y = slope_educ)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.8) +
  geom_col(aes(fill = slope_educ < 0), alpha = 0.8, width = 0.6) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0.2, linewidth = 0.9) +
  geom_text(aes(label = sprintf("%.2f", slope_educ),
                y = slope_educ + sign(slope_educ)*0.03),
            size = 3.5, fontface = "bold") +
  scale_fill_manual(values = c("TRUE" = "#0073C2FF", "FALSE" = "#F8766D"),
                    labels = c("Civilizing effect (reduces)", "Increases justification"),
                    name = "Direction of effect") +
  labs(
    title = "Civilizing Effect of Education among Non-Participants",
    subtitle = "Negative slopes confirm that higher education reduces violence justification",
    x = "Education Level",
    y = "Difference in Justification\n(vs. High School or less)",
    caption = "Bars show differences relative to secondary education; vertical lines: 95% CI. Dashed line at 0: no difference. Compiled by the author based on ELSOC 2016–2023."
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1, size = 7),
    legend.position = "bottom",
    legend.text = element_text(size = 7),
    plot.title = element_text(face = "bold", size = 9),
    plot.subtitle = element_text(size = 7.5, color = "gray30"),
    plot.caption = element_text(size = 6.5, hjust = 0)
  )

  slopes_no_part
```

@fig-slopes-educacion-nopart visualizes the classic civilizing effect among those who do not participate in protests, showing how each higher education level reduces violence justification compared to the reference category (High School or less). The negative bars (blue) confirm that all technical and university education levels are associated with lower justification among external observers, although with varying magnitudes. This pattern contrasts radically with the effect among participants (@fig-slopes-educacion), where education amplifies justification, evidencing the contextual dependence of cultural capital.

@fig-slopes-educacion complements the crossed-lines plot by visualizing the magnitude of the observed participation gap at each education level — that is, the difference in mean violence justification between those who participated and those who did not. These slopes capture a combination of within-person socialization and between-person selection; the within-between analysis in Section 5.2.1 suggests that the between-person (sorting) component dominates.

Among those with secondary education or less, the observed gap is +0.15 (CI: [0.05, 0.25]), consistent with a modest selection effect: those who participate from this educational background differ only slightly in mean justification from non-participants with the same education. The gap increases at higher education levels: technical college slopes range from +0.25 to +0.35, and university complete reaches +0.35 (CI: [0.22, 0.48]). The wider gap at higher education levels reflects the stronger sorting documented in the between-person component of the REWB interaction models: university-educated habitual protesters are more attitudinally distinctive from their non-participating peers than lower-educated protesters are from theirs.

One notable feature of @fig-slopes-educacion is that the largest single gap appears at technical college incomplete rather than at university complete. This may reflect the small and potentially non-representative subsample of individuals in this category who participate in protests, or compositional differences (e.g., age, occupation, or sector of enrollment) that make this group disproportionately selected. The interaction term for this group is imprecisely estimated and should be interpreted with caution; it does not alter the general monotonic pattern across the remaining education levels.

@fig-paradoja-educacion shows complex patterns of educational moderation on protest violence justification.

Among those who have not participated in protests, most higher education levels show significant negative effects compared to complete secondary education or less (reference category). Technical college incomplete reduces justification by -0.13 (p<0.05), technical complete by -0.13 (p<0.001), and university complete by -0.13 (p<0.001). University incomplete shows a smaller and non-significant effect (-0.07). This is consistent with a "civilizing effect" of technical and complete university education: higher credentialing is associated with lower violence justification among external observers.

Protest participation significantly increases violence justification ($+0.27$, p<0.001) for the reference group (secondary education or less), a substantial and meaningful effect.

Regarding interaction terms, all are positive, suggesting that higher education magnifies the association between participation and justification, though with different magnitudes. Technical college incomplete shows the strongest interaction ($+0.28$, p<0.05), followed by technical complete and university complete (both $+0.13$, p<0.05 and p<0.01 respectively), while university incomplete does not reach statistical significance ($+0.11$, ns). Calculating the total effect for complete university participants: $-0.13$ (main effect) $+0.27$ (participation) $+0.13$ (interaction) = $+0.27$. This implies that university completers who protest justify violence at the same level as the reference group with secondary education, completely nullifying the initial civilizing effect.

The year 2019 shows a positive coefficient ($+0.05$, p<0.05), consistent with the social outbreak context. This suggests that, controlling for education and participation, 2019 had relatively higher justification. Other post-2016 years show negative coefficients (2018: $-0.11***$; 2022 and 2023: $-0.10***$), indicating a general downward trend, while 2017 does not differ significantly from the base year.

Finally, controls show that age reduces justification ($-0.01$, p<0.001), right-wing ideology also reduces it ($-0.03$, p<0.001), while gender (woman) has no significant effect ($-0.03$, ns). The model shows a random intercept variance of 0.13, indicating substantial individual heterogeneity in baseline justification levels.
